---
title: "Allocate: Assigning Intensity Bands"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Allocate: Assigning Intensity Bands}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4,
  warning  = FALSE,
  message  = FALSE
)
```

```{r pipeline-setup, include=FALSE}
library(motiongrammar)
library(ggplot2)
library(dplyr)

# Build through designate — we demonstrate allocate() here
trace_pre <- initiate(
  session = system.file("extdata", "example_gps_file_short.csv", package = "motiongrammar"),
  source  = 'guess_csv'
) |>
  coordinate(from = "latlong", to = "xyz", crs = 3857, norm = TRUE,
             drop_outliers = TRUE) |>
  interpolate(hz = 10, max_gap_frames = 20) |>
  filtrate(method = "butterworth", target = "coordinates", cutoff = 1.5) |>
  derivate(derivatives = c("velocity", "acceleration", "angular_velocity")) |>
  designate(method = "corbett", v_threshold = 0.3)
```

`allocate()` cuts each derivative column into labelled intensity bands
using user-defined thresholds. The result is a new ordered-factor column
for each derivative that `quantitate()` uses to produce load profiles.

------------------------------------------------------------------------

## 1. Why intensity bands?

Raw velocity tells you what happened in each frame. However, ultimately,
we want to be able to answer questions such as:

1.  How much of a players session was spent walking?
2.  How much did a player spring during a session?

To answer these questions, we set thresholds, and calculate things like
efforts and total distances within these allocations. By specifying the
same thresholds for every player and every session, you create a
consistent, comparable taxonomy.

Eventually, I'll add support for *Sweeting's Clustering Method* to
actually determine where the cut off lies. But for version 1 of
motiongrammar, the goal is to apply bands identified elsewhere.

------------------------------------------------------------------------

## 2. Basic usage

```{r allocate}
library(motiongrammar)
library(ggplot2)
library(dplyr)

trace_pre <- initiate(
  session = system.file("extdata", "example_gps_file_short.csv", package = "motiongrammar"),
  source  = 'guess_csv'
) |>
  coordinate(from = "latlong", to = "xyz", crs = 3857, norm = TRUE,
             drop_outliers = TRUE) |>
  interpolate(hz = 10, max_gap_frames = 20) |>
  filtrate(method = "butterworth", target = "coordinates", cutoff = 1.5) |>
  derivate(derivatives = c("velocity", "acceleration", "angular_velocity")) |>
  designate(method = "corbett", v_threshold = 0.3)

trace <- trace_pre |>
  allocate(
    allocation = list(
      velocity     = c(0, 2, 4, 6, 8),
      acceleration = c(0, 1, 2, 3, 4),
      angular_velocity = c(0, 90, 180, 270)
    ),
    allocation_name = "gps_standard"
  )

# New columns added:
grep("band", names(trace), value = TRUE)
```

------------------------------------------------------------------------

## 3. How thresholds define bands

A vector of N values creates N−1 bands:

```         
velocity = c(0, 2, 4, 6, 8)
```

| Band     | Range     | Arbitrary Label |
|----------|-----------|-----------------|
| `band_1` | 0 – 2 m/s | Walk            |
| `band_2` | 2 – 4 m/s | Jog             |
| `band_3` | 4 – 6 m/s | Run             |
| `band_4` | 6+ m/s    | High-speed run  |

Values above the highest threshold fall into the top band (open upper
bound). Values below the lowest threshold become `NA`. Bands are stored
as **ordered factors** so `band_1 < band_2 < band_3` — comparisons and
bar chart ordering work correctly.

Acceleration and angular velocity use `abs()` before banding — a
deceleration of −4 m/s² and an acceleration of +4 m/s² represent equal
magnitudes of loading.

------------------------------------------------------------------------

## 4. Visualising the band distribution

```{r band-bar}
# Time in each velocity band across the active segment
trace |>
  filter(designation == "active_1", !is.na(gps_standard_velocity_band)) |>
  count(gps_standard_velocity_band) |>
  mutate(duration_s = n / 10) |>
  ggplot(aes(x = gps_standard_velocity_band, y = duration_s,
             fill = gps_standard_velocity_band)) +
  geom_col(show.legend = FALSE) +
  scale_fill_viridis_d(option = "plasma", begin = 0.1, end = 0.9) +
  labs(
    title = "Time in each velocity band (active segment)",
    x     = "Velocity band",
    y     = "Duration (s)"
  ) +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

```{r band-traj, fig.height=5}
# Colour the trajectory by velocity band
trace |>
  filter(!is.na(f_x), !is.na(gps_standard_velocity_band)) |>
  ggplot(aes(x = f_x, y = f_y, colour = gps_standard_velocity_band)) +
  geom_path(linewidth = 1.2) +
  coord_fixed() +
  scale_colour_viridis_d(option = "plasma", begin = 0.1, end = 0.9) +
  labs(
    title  = "Trajectory coloured by velocity band",
    x      = "x (m)", y = "y (m)",
    colour = "Velocity\nband"
  ) +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

------------------------------------------------------------------------

## 5. Multiple allocation schemes

Call `allocate()` multiple times with different names to layer multiple
profiling schemes onto the same trace:

```{r multiple}
trace <- trace_pre |>
  allocate(
    allocation      = list(velocity = c(0, 2, 4, 6, 8)),
    allocation_name = "club_standard"
  ) |>
  allocate(
    allocation      = list(velocity = c(0, 1.5, 3, 5, 6.5)),
    allocation_name = "literature_2023"
  )

grep("band", names(trace), value = TRUE)
```

------------------------------------------------------------------------

## 6. Loading thresholds from a CSV — `allocate_csv()`

For team environments where thresholds are managed centrally, store them
in a CSV and load with `allocate_csv()`:

```{r allocate-csv, eval=FALSE}
trace <- trace_pre |>
  allocate_csv(file = "~/config/thresholds.csv")
```

**CSV format:**

``` csv
profile_name,target_derivative,profile_start,profile_end
gps_standard,velocity,0,2
gps_standard,velocity,2,4
gps_standard,velocity,4,6
gps_standard,velocity,6,8
gps_standard,acceleration,0,1
gps_standard,acceleration,1,2
gps_standard,acceleration,2,3
gps_standard,acceleration,3,4
```

The source file path is stored in the quality log, so provenance is
maintained even when thresholds change between seasons.

------------------------------------------------------------------------

## 7. Reading the quality log

```{r quality}
quality_report(trace, step = 'allocate')
```

The allocate section records each named scheme separately: band count,
threshold values, the range of each band, and whether the source was
inline code or a CSV file.

------------------------------------------------------------------------

## Quick reference

| Argument          | Default          | Purpose                            |
|-------------------|------------------|------------------------------------|
| `allocation`      | Standard GPS set | Named list of threshold vectors    |
| `allocation_name` | `"allocation_1"` | Prefix for the output band columns |

`allocate_csv()` — `file` is the only argument: path to a CSV with
`profile_name`, `target_derivative`, `profile_start`, `profile_end`.
