---
title: "Coordinate: Projecting GPS to Cartesian Space"
output: rmarkdown::html_vignette
vignette: >
%\VignetteIndexEntry{Coordinate: Projecting GPS to Cartesian Space}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
  editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4,
  warning  = FALSE,
  message  = FALSE
)
```

```{r load-pkg, include=FALSE}
devtools::load_all()
library(ggplot2)
library(dplyr)

# Load raw GPS trace — we demonstrate coordinate() in this vignette
trace_raw <- initiate(
  session = '~/motiongrammar/data/example_gps_file_short.csv',
  source  = 'catapult_replay'
)
```

`coordinate()` is the second step in every GPS pipeline. It converts the
raw latitude and longitude values produced by `initiate()` into a flat,
metre-scale Cartesian plane where distances, velocities, and headings
become meaningful numbers.

------------------------------------------------------------------------

## 1. Why projection matters

GPS coordinates are angles on a sphere (degrees of latitude and
longitude). You cannot compute Euclidean distance, velocity, or
acceleration directly from them. A one-degree change in longitude means
roughly 111 km at the equator but only 78 km at 45° latitude — and the
difference grows as you move toward the poles.

`coordinate()` projects those angles onto a flat 2-D surface using a
standard coordinate reference system (CRS), so that every unit is
metres and arithmetic is geometrically correct.

------------------------------------------------------------------------

## 2. Basic usage

```{r coordinate-basic}

# Load in our key packages, and initiate the test file
library(motiongrammar)
library(ggplot2)
library(dplyr)

trace_raw <- initiate(session = '~/motiongrammar/data/example_gps_file_short.csv',
                      source  = 'catapult_replay')

trace <- trace_raw |>
  coordinate(from = "latlong", 
             to = "xyz", 
             crs = 3857)
```

After `coordinate()` the trace gains three new columns:

| Column | Content                                               |
|--------|-------------------------------------------------------|
| `x`    | Easting in metres (positive = East)                   |
| `y`    | Northing in metres (positive = North)                 |
| `z`    | Altitude in metres (from `altitude` if present, otherwise 0) |

The original `lat`, `lng`, and `altitude` columns are kept alongside
them so you can always trace back to the raw values.

```{r plot-raw}

trace |>
  filter(!is.na(x)) |>
  ggplot(aes(x = x, y = y)) +
  geom_path(alpha = 0.5, linewidth = 0.3) +
  coord_fixed() +
  labs(
    title = "Projected trajectory (absolute coordinates)",
    x = "Easting (m — EPSG:3857)",
    y = "Northing (m — EPSG:3857)"
  ) +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

The axes are six-to-eight digit numbers — correct but hard to interpret.
`norm = TRUE` fixes this.

------------------------------------------------------------------------

## 3. Choosing a CRS

The CRS (coordinate reference system) controls how degrees are projected
into metres.

| EPSG code | Name                         | Best for                          |
|-----------|------------------------------|-----------------------------------|
| **3857**  | Web Mercator (default)       | Most sports science applications  |
| **32754** | UTM Zone 54S                 | Australia / New Zealand           |
| **27700** | British National Grid        | United Kingdom                    |
| **2263**  | New York State Plane         | Specific US regions               |

For field-sport GPS the differences between projections are sub-centimetre
over a playing arena. Web Mercator (`3857`) is the safe default.

------------------------------------------------------------------------

## 4. Normalisation — anchoring the origin

### `norm = TRUE` — first valid position becomes (0, 0, 0)

```{r norm-true}
trace <- trace_raw |>
  coordinate(from = "latlong", 
             to = "xyz", 
             crs = 3857, 
             norm = TRUE)
```

```{r plot-normed}
trace |>
  filter(!is.na(x)) |>
  ggplot(aes(x = x, y = y)) +
  geom_path(alpha = 0.5, linewidth = 0.4, colour = "steelblue") +
  coord_fixed() +
  labs(
    title = "Projected trajectory — normalised to first position",
    subtitle = "Origin (0, 0) is where the player first appeared in the data",
    x = "x (m)",
    y = "y (m)"
  ) +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

Every subsequent position is expressed as metres from where the player
was first recorded. Plots are compact and comparable across sessions.

### `norm = c(x, y)` — custom absolute origin

If you want all players in a session to share a common reference point
(e.g. the south-west corner of the field), pass an absolute projected
coordinate. All players will share the same (0, 0), so trajectories can
be overlaid on a field diagram.

```{r norm-custom, eval=FALSE}
field_sw_corner <- c(-13952.6,
                     6711140.2)

trace <- trace_raw |>
  coordinate(
    from = "latlong",
    to   = "xyz",
    crs  = 3857,
    norm = field_sw_corner
  )
```

### `norm = FALSE` — keep absolute projected coordinates

Use this when you need to do spatial joins with other geographic data
or overlay on a map tile.

------------------------------------------------------------------------

## 5. Rotating to align with the field

Field sport data is almost never recorded with the long axis of the
pitch aligned to the x-axis. The `rotate` argument takes two GPS points
defining a sideline and rotates the entire coordinate system so that
sideline lies along x (and the perpendicular along y).

```{r rotate, eval=FALSE}
sideline_sw <- c(-35.318, 149.133)
sideline_se <- c(-35.318, 149.136)

trace <- trace_raw |>
  coordinate(
    from   = "latlong",
    to     = "xyz",
    crs    = 3857,
    norm   = TRUE,
    rotate = list(sideline_sw, sideline_se)
  )
```

After rotation, x increases along the long axis of the field and y
increases toward the far touchline. The rotation angle (radians and
degrees) is recorded in both `metadata` and the quality log.

> **Tip:** Collect two sideline GPS points from Google Maps before the
> session. Any two points along the same line are sufficient — they do
> not need to be the exact corners.

------------------------------------------------------------------------

## 6. Dropping outliers

GPS occasionally produces errant readings — phantom positions hundreds of
metres off-field. `drop_outliers = TRUE` replaces any position outside
**3 × IQR** fences on x or y with `NA`.

```{r drop-outliers}
trace <- trace_raw |>
  coordinate(
    from          = "latlong",
    to            = "xyz",
    crs           = 3857,
    norm          = TRUE,
    drop_outliers = TRUE
  )
```

```{r plot-outliers}
trace |>
  filter(!is.na(x)) |>
  ggplot(aes(x = x, y = y)) +
  geom_path(alpha = 0.5, linewidth = 0.4, colour = "steelblue") +
  coord_fixed() +
  labs(
    title = "Trajectory after outlier removal",
    subtitle = "Points outside 3 × IQR set to NA; interpolate() fills them next",
    x = "x (m)", y = "y (m)"
  ) +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

The outlier count, percentage, and IQR bounds are recorded in the quality
log. `interpolate()` will fill the resulting NAs in the next step.

------------------------------------------------------------------------

## 7. Unit conversion for local-coordinate data

If your data is already in local Cartesian coordinates (e.g. yards from
a local positioning system), `coordinate()` can convert units in the same
step without any reprojection:

```{r units, eval=FALSE}
trace <- initiate(
  session  = "lps_session.csv",
  source   = "manual_csv",
  col_unix = "Time",
  col_lat  = "X_yd",
  col_lng  = "Y_yd"
) |>
  coordinate(
    from       = "xyz",
    to         = "xyz",
    from_units = "yd",
    to_units   = "m"
  )
```

Supported unit strings: `"mm"`, `"cm"`, `"m"`, `"km"`, `"in"`, `"ft"`,
`"yd"`, `"mi"`.

------------------------------------------------------------------------

## 8. Reading the quality log

```{r quality}
quality_report(trace)
```

The coordinate section reports the conversion path, CRS codes and name,
normalisation origin, rotation angle if applied, x and y range in
metres, zero-GPS rows removed, and any outliers flagged or dropped.

------------------------------------------------------------------------

## Quick reference

| Argument        | Default      | Purpose                                           |
|-----------------|--------------|---------------------------------------------------|
| `from`          | `"latlong"`  | Source coordinate type                            |
| `to`            | `"xyz"`      | Target type (always `"xyz"`)                      |
| `crs`           | `3857`       | EPSG projection code                              |
| `norm`          | `FALSE`      | `TRUE` → first-row origin; `c(x,y)` → custom     |
| `rotate`        | `NULL`       | List of two `c(lat, lng)` sideline points         |
| `drop_outliers` | `FALSE`      | Replace 3×IQR outliers with NA                    |
| `from_units`    | `NULL`       | Source units for xyz → xyz conversion             |
| `to_units`      | `"m"`        | Target units                                      |
