---
title: "Derivate: Computing Velocity, Acceleration, and Beyond"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Derivate: Computing Velocity, Acceleration, and Beyond}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4,
  warning  = FALSE,
  message  = FALSE
)
```

```{r pipeline-setup, include=FALSE}
devtools::load_all()
library(ggplot2)
library(dplyr)
library(tidyr)

# Build through filtrate — we demonstrate derivate() here
trace_pre <- initiate(
  session = '~/motiongrammar/data/example_gps_file_short.csv',
  source  = 'catapult_replay'
) |>
  coordinate(from = "latlong", to = "xyz", crs = 3857, norm = TRUE,
             drop_outliers = TRUE) |>
  interpolate(hz = 10, max_gap_frames = 20) |>
  filtrate(method = "butterworth", target = "coordinates", cutoff = 1.5)
```

`derivate()` computes time-series derivatives of a projected and filtered
motion trace. It always produces `distance` and `heading`, and adds
whichever of velocity, acceleration, angular velocity, jerk, and lateral
G you request.

------------------------------------------------------------------------

## 1. Basic usage

```{r basic}
library(motiongrammar)
library(ggplot2)
library(dplyr)
library(tidyr)

trace_pre <- initiate(
  session = '~/motiongrammar/data/example_gps_file_short.csv',
  source  = 'catapult_replay'
) |>
  coordinate(from = "latlong", to = "xyz", crs = 3857, norm = TRUE,
             drop_outliers = TRUE) |>
  interpolate(hz = 10, max_gap_frames = 20) |>
  filtrate(method = "butterworth", target = "coordinates", cutoff = 1.5)

trace <- trace_pre |>
  derivate(derivatives = c("velocity", "acceleration", "angular_velocity"))

names(trace)
summary(trace[, c("velocity", "acceleration", "angular_velocity")])
```

------------------------------------------------------------------------

## 2. Velocity

Velocity is Euclidean displacement divided by elapsed time over the
preceding window:

```
velocity = sqrt((x[i] - x[i-w])² + (y[i] - y[i-w])²) / (t[i] - t[i-w])
```

```{r velocity-ts}
# Plot velocity time-series (only the GPS period)
trace |>
  filter(!is.na(f_x)) |>
  ggplot(aes(x = unix_time, y = velocity)) +
  geom_line(linewidth = 0.3, colour = "steelblue") +
  labs(
    title = "Velocity over time",
    x     = "Unix time (s)",
    y     = "Speed (m/s)"
  ) +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

------------------------------------------------------------------------

## 3. Acceleration

```{r acceleration-ts}
trace |>
  filter(!is.na(f_x)) |>
  ggplot(aes(x = unix_time, y = acceleration)) +
  geom_line(linewidth = 0.3, colour = "darkorange") +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  labs(
    title = "Acceleration over time",
    x     = "Unix time (s)",
    y     = "Acceleration (m/s²)"
  ) +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

Typical ranges in field sport:

| Context              | Range            |
|----------------------|------------------|
| Normal running       | −3 to +3 m/s²    |
| High-intensity burst | ±4 to ±6 m/s²    |
| Artefact (noise)     | > ±10 m/s²       |

------------------------------------------------------------------------

## 4. Angular velocity — measuring curvature

Angular velocity is derived from heading using a finite difference with
circular wrapping to handle the 360° → 0° boundary:

```
Δh = ((heading[i] - heading[i-w] + 180) %% 360) - 180
angular_velocity = |Δh| / dt
```

Zero when running straight; high when changing direction sharply.

```{r angular-ts}
trace |>
  filter(!is.na(f_x)) |>
  ggplot(aes(x = unix_time, y = angular_velocity)) +
  geom_line(linewidth = 0.3, colour = "mediumpurple") +
  labs(
    title = "Angular velocity over time",
    x     = "Unix time (s)",
    y     = "Angular velocity (°/s)"
  ) +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

### Angular velocity on the field trajectory

Colouring the path by angular velocity reveals where the player turned:

```{r angular-field, fig.height=5}
trace |>
  filter(!is.na(f_x), angular_velocity < 200) |>
  ggplot(aes(x = f_x, y = f_y, colour = angular_velocity)) +
  geom_path(linewidth = 1.2) +
  coord_fixed() +
  scale_colour_viridis_c(option = "plasma", begin = 0.1, end = 0.9) +
  labs(
    title  = "Trajectory coloured by angular velocity",
    subtitle = "Bright = sharp turn; dark = straight running",
    x      = "x (m)", y = "y (m)",
    colour = "°/s"
  ) +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

------------------------------------------------------------------------

## 5. All derivatives together

```{r all-derivs, fig.height=7}
trace |>
  filter(!is.na(f_x)) |>
  select(unix_time, velocity, acceleration, angular_velocity) |>
  pivot_longer(-unix_time, names_to = "derivative", values_to = "value") |>
  mutate(derivative = factor(derivative,
    levels = c("velocity", "acceleration", "angular_velocity"),
    labels = c("Velocity (m/s)", "Acceleration (m/s²)", "Angular velocity (°/s)")
  )) |>
  ggplot(aes(x = unix_time, y = value)) +
  geom_line(linewidth = 0.3, colour = "steelblue") +
  facet_wrap(~ derivative, ncol = 1, scales = "free_y") +
  labs(title = "All derivatives", x = "Unix time (s)", y = NULL) +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

------------------------------------------------------------------------

## 6. The `window` argument

The window controls how many frames span each finite difference. Larger
windows average over more time (less noise, less temporal resolution).

```{r window, eval=FALSE}
# Uniform window: 5 frames at 10 Hz = 0.5 s differences
trace <- trace_pre |> derivate(window = 5)

# Per-derivative windows: velocity wider, acceleration narrower
trace <- trace_pre |>
  derivate(
    window = list(
      velocity         = 10,    # 1 s at 10 Hz
      acceleration     =  5,    # 0.5 s
      angular_velocity = 10
    )
  )
```

The default window equals the sample rate in Hz (10 frames at 10 Hz = 1-second differences),
making the default consistent across different device sample rates.

------------------------------------------------------------------------

## 7. `use_filtered` and `speed_floor`

```{r options, eval=FALSE}
# Force raw coordinates — not recommended without a reason
trace <- trace_pre |> derivate(use_filtered = FALSE)

# Suppress heading below 0.2 m/s (very slow walk)
trace <- trace_pre |> derivate(speed_floor = 0.2)
```

Below `speed_floor`, `heading` and `angular_velocity` are suppressed to
avoid noise in direction measurements when the player is stationary.

------------------------------------------------------------------------

## 8. Signed angular velocity

```{r signed-omega, eval=FALSE}
# Retain sign: positive = left / counter-clockwise
trace <- trace_pre |>
  derivate(derivatives = "angular_velocity", signed_omega = TRUE)
```

Useful for quantifying movement asymmetry — e.g. more left turns than
right in a particular drill pattern.

------------------------------------------------------------------------

## 9. Experimental: jerk and lateral G

```{r experimental, eval=FALSE}
# Third-order derivative — very sensitive to noise, use with caution
trace <- trace_pre |>
  derivate(derivatives = c("velocity", "acceleration", "angular_velocity", "jerk"))

# Centripetal load proxy (v × ω_rad / g)
trace <- trace_pre |>
  derivate(derivatives = "all")
```

Both print a warning. Jerk and lateral G amplify GPS positional noise
significantly and have limited validation in the sports science
literature. Apply aggressive coordinate filtering (cutoff ≤ 1 Hz) if
you need them.

------------------------------------------------------------------------

## 10. Reading the quality log

```{r quality}
quality_report(trace)
```

The derivate section records which coordinate source was used
(`f_x`/`f_y` or raw), all window sizes, `speed_floor`, and
`signed_omega`. It also logs the full algorithm formula for each
derivative — copy these into a methods section for a publication.

------------------------------------------------------------------------

## Quick reference

| Argument       | Default     | Purpose                                                      |
|----------------|-------------|--------------------------------------------------------------|
| `use_filtered` | `TRUE`      | Use `f_x`/`f_y` if available; fall back to raw with message  |
| `derivatives`  | standard set | `"velocity"`, `"acceleration"`, `"angular_velocity"`, `"jerk"`, `"lateral_g"`, or `"all"` |
| `window`       | sample rate Hz | Frames per finite difference; scalar or named list         |
| `speed_floor`  | `0.5`       | m/s below which heading / angular velocity are suppressed    |
| `signed_omega` | `FALSE`     | Retain sign in angular velocity (+ = CCW, − = CW)           |
