---
title: "Derivate: Computing Velocity, Acceleration, and Beyond"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Derivate: Computing Velocity, Acceleration, and Beyond}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4,
  warning  = FALSE,
  message  = FALSE
)
```

```{r pipeline-setup, include=FALSE}
devtools::load_all()
library(ggplot2)
library(dplyr)
library(tidyr)

# Build through filtrate — we demonstrate derivate() here
trace_pre <- initiate(
  session = '~/motiongrammar/data/example_gps_file_short.csv',
  source  = 'catapult_replay'
) |>
  coordinate(from = "latlong", to = "xyz", crs = 3857, norm = TRUE,
             drop_outliers = TRUE) |>
  interpolate(hz = 10, max_gap_frames = 20) |>
  filtrate(method = "butterworth", target = "coordinates", cutoff = 1.5)
```

`derivate()` computes time-series derivatives of a projected and
filtered motion trace. It always produces `distance` and `heading`, and
adds whichever of velocity, acceleration, angular velocity, jerk, and
lateral G you request.

------------------------------------------------------------------------

## 1. Basic usage

```{r basic}
library(motiongrammar)
library(ggplot2)
library(dplyr)
library(tidyr)

trace_pre <- initiate(
  session = '~/motiongrammar/data/example_gps_file_short.csv',
  source  = 'catapult_replay'
) |>
  coordinate(from = "latlong", to = "xyz", crs = 3857, norm = TRUE,
             drop_outliers = TRUE) |>
  interpolate(hz = 10, max_gap_frames = 20) |>
  filtrate(method = "butterworth", target = "coordinates", cutoff = 1.5)

trace <- trace_pre |>
  derivate(derivatives = c("velocity", "acceleration", "angular_velocity"))

names(trace)
summary(trace[, c("velocity", "acceleration", "angular_velocity")])
```

------------------------------------------------------------------------

## 2. Velocity

Velocity measures how fast a player is going, without accounting for
things like change of direction etc. Scientifically, velocity and speed
(and distance/displacement) are not interchangeable. However, velocity
is colloquially the most common term used in sport to describe "how
fast" a player is running, and is therefore what I use here.

*motiongrammar* computes velocity using the formula below.

```         
velocity = sqrt((x[i] - x[i-w])² + (y[i] - y[i-w])²) / (t[i] - t[i-w])
```

When we look at our velocity visuals below, we are asking *how fast is a
player, over time*? Higher time points (like our erroneous spike at the
end) denote faster time points. Whereas points closer to 0, mean a
player is closer to being stationary.

```{r velocity-ts}
# Plot velocity time-series (only the GPS period)
trace |>
  filter(!is.na(f_x)) |>
  ggplot(aes(x = unix_time, y = velocity)) +
  geom_line(linewidth = 0.5, colour = "black") +
  labs(
    title = "Velocity over time",
    x     = "Unix time (s)",
    y     = "Velocity (m/s)"
  ) +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

------------------------------------------------------------------------

## 3. Acceleration

Acceleration measures how quickly a player *speeds up* or *slows down*.
From a motion perspective, we consider acceleration as a single
time-series. Positive values mean a player is getting faster (ie., their
current velocity is higher than their previous velocity). Negative
values mean a player is decelerating (ie., their current velocity is
lower than their previous velocity).

Please note, that acceleration is notoriously noisy in **any setting**.
This is especially true using relatively low fidelity GPS data (as
opposed to an optical system like Hawkeye). My advice is to tread
carefully when analysing acceleration. It's common in the field and many
practitioners do it, but you are dealing with a lot more noise here.

```{r acceleration-ts}
trace |>
  filter(!is.na(f_x)) |>
  ggplot(aes(x = unix_time, y = acceleration)) +
  geom_line(linewidth = 0.3, colour = "darkorange") +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  labs(
    title = "Acceleration over time",
    x     = "Unix time (s)",
    y     = "Acceleration (m/s²)"
  ) +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

## 4. Angular velocity

We use angular velocity to get a time-series of "change of direction".
Both velocity and acceleration measure how fast a player moves. Angular
velocity, on the other hand, is derived from heading using a finite
difference with circular wrapping to handle the 360° → 0° boundary:

```         
Δh = ((heading[i] - heading[i-w] + 180) %% 360) - 180
angular_velocity = |Δh| / dt
```

Zero when running straight; high when changing direction sharply.

```{r angular-ts}
trace |>
  filter(!is.na(f_x)) |>
  ggplot(aes(x = unix_time, y = angular_velocity)) +
  geom_line(linewidth = 0.3, colour = "mediumpurple") +
  labs(
    title = "Angular velocity over time",
    x     = "Unix time (s)",
    y     = "Angular velocity (°/s)"
  ) +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

### Angular velocity on the field trajectory

Colouring the path by angular velocity reveals where the player turned:

```{r angular-field, fig.height=5}
trace |>
  filter(!is.na(f_x), angular_velocity < 200) |>
  ggplot(aes(x = f_x, y = f_y, colour = angular_velocity)) +
  geom_path(linewidth = 1.2) +
  coord_fixed() +
  scale_colour_viridis_c(option = "plasma", begin = 0.1, end = 0.9) +
  labs(
    title  = "Trajectory coloured by angular velocity",
    subtitle = "Bright = sharp turn; dark = straight running",
    x      = "x (m)", y = "y (m)",
    colour = "°/s"
  ) +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

------------------------------------------------------------------------

## 5. All derivatives together

```{r all-derivs, fig.height=7}
trace |>
  filter(!is.na(f_x)) |>
  select(unix_time, velocity, acceleration, angular_velocity) |>
  pivot_longer(-unix_time, names_to = "derivative", values_to = "value") |>
  mutate(derivative = factor(derivative,
    levels = c("velocity", "acceleration", "angular_velocity"),
    labels = c("Velocity (m/s)", "Acceleration (m/s²)", "Angular velocity (°/s)")
  )) |>
  ggplot(aes(x = unix_time, y = value)) +
  geom_line(linewidth = 0.3, colour = "steelblue") +
  facet_wrap(~ derivative, ncol = 1, scales = "free_y") +
  labs(title = "All derivatives", x = "Unix time (s)", y = NULL) +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

------------------------------------------------------------------------

## 6. The `window` argument

We do not usually calculate *truly instantaneous* velocity, acceleration
and angular velocity. Instead, we usually look within *a window*. This
adds an additional check/balance to single data points where a players
acceleration or velocity may unexpectedly spike even after filtering.

The window argument controls how many frames span each finite
difference. Larger windows average over more time (less noise, less
temporal resolution). We can set the window globally (for example window
= 5), or supply different windows for velocity, acceleration and
angular_velocity specifically.

```{r window, eval=FALSE}
# Uniform window: 5 frames at 10 Hz = 0.5 s differences
trace <- trace_pre |> derivate(window = 5)

# Per-derivative windows: velocity wider, acceleration narrower
trace <- trace_pre |>
  derivate(
    window = list(
      velocity         = 10,    # 1 s at 10 Hz
      acceleration     =  5,    # 0.5 s
      angular_velocity = 10
    )
  )
```

------------------------------------------------------------------------

## 7. `use_filtered` and `speed_floor`

```{r options, eval=FALSE}
# Force raw coordinates — not recommended without a reason
trace <- trace_pre |> derivate(use_filtered = FALSE)

# Suppress heading below 0.2 m/s (very slow walk)
trace <- trace_pre |> derivate(speed_floor = 0.2)
```

use-filtered tells R whether you want to use filtered data from a
previous step, or raw data. If you are working with already filtered
data from a provider, it makes sense the set use_filtered as FALSE.

Below `speed_floor`, `heading` and `angular_velocity` are suppressed to
avoid noise in direction measurements when the player is stationary.

------------------------------------------------------------------------

## 8. Signed angular velocity

Signed omega allows us to determine the direction a player turns/changes
direction. I'm sure there is someone really smart out there who could
use this to identify asymmetries in load using this, so I included it
here.

```{r signed-omega, eval=FALSE}
# Retain sign: positive = left / counter-clockwise
trace <- trace_pre |>
  derivate(derivatives = "angular_velocity", signed_omega = TRUE)
```

------------------------------------------------------------------------

## 9. Experimental: jerk and lateral G

```{r experimental, eval=FALSE}
# Third-order derivative — very sensitive to noise, use with caution
trace <- trace_pre |>
  derivate(derivatives = c("velocity", "acceleration", "angular_velocity", "jerk"))

# Centripetal load proxy (v × ω_rad / g)
trace <- trace_pre |>
  derivate(derivatives = "all")
```

Both print a warning. Jerk and lateral G amplify GPS positional noise
significantly and have limited validation in the sports science
literature. I would not use these.

------------------------------------------------------------------------

## 10. Reading the quality log

```{r quality}
quality_report(trace)
```

The derivate section records which coordinate source was used
(`f_x`/`f_y` or raw), all window sizes, `speed_floor`, and
`signed_omega`. It also logs the full algorithm formula for each
derivative — copy these into a methods section for a publication.

------------------------------------------------------------------------

## Quick reference

| Argument | Default | Purpose |
|--------------|--------------|--------------------------------------------|
| `use_filtered` | `TRUE` | Use `f_x`/`f_y` if available; fall back to raw with message |
| `derivatives` | standard set | `"velocity"`, `"acceleration"`, `"angular_velocity"`, `"jerk"`, `"lateral_g"`, or `"all"` |
| `window` | sample rate Hz | Frames per finite difference; scalar or named list |
| `speed_floor` | `0.5` | m/s below which heading / angular velocity are suppressed |
| `signed_omega` | `FALSE` | Retain sign in angular velocity (+ = CCW, − = CW) |
