---
title: "Designate: Segmenting a Session into Episodes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Designate: Segmenting a Session into Episodes}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4,
  warning  = FALSE,
  message  = FALSE
)
```

```{r pipeline-setup, include=FALSE}
devtools::load_all()
library(ggplot2)
library(dplyr)

# Build through derivate — we demonstrate designate() here
trace_pre <- initiate(
  session = '~/motiongrammar/data/example_gps_file_short.csv',
  source  = 'catapult_replay'
) |>
  coordinate(from = "latlong", to = "xyz", crs = 3857, norm = TRUE,
             drop_outliers = TRUE) |>
  interpolate(hz = 10, max_gap_frames = 20) |>
  filtrate(method = "butterworth", target = "coordinates", cutoff = 1.5) |>
  derivate(derivatives = c("velocity", "acceleration", "angular_velocity"))
```

`designate()` partitions a continuous motion trace into labelled
"sections", active bouts of movement, drills or game intervals, and
relief periods. Quantitate can summarise each section separately.

------------------------------------------------------------------------

## 1. Why we designate

A GPS file for a training session is one long continuous time-series.
Aggregating the whole file without segmentation conflates a warm-up,
multiple drill blocks, and a cool-down into a single number.
`designate()` adds a `designation` column so each episode gets its own
summary row.

Chances are, you have designations (for example drill names) in your
providers file. That's great! If that's the case, it makes total sense
to preserve those columns at the *initiate* phase and group by them.
Designate is built with automatic detection in mind (using a variation
on my PhD methods), as well as for instances where labels may be stored
in a separate csv. Drills are the most obvious thing we would designate
by, but you could also use this to look at any tagged event or context
from Sportscode.

------------------------------------------------------------------------

## 2. Automatic segmentation — `method = 'corbett'`

```{r designate}
library(motiongrammar)
library(ggplot2)
library(dplyr)

trace_pre <- initiate(
  session = '~/motiongrammar/data/example_gps_file_short.csv',
  source  = 'catapult_replay'
) |>
  coordinate(from = "latlong", to = "xyz", crs = 3857, norm = TRUE,
             drop_outliers = TRUE) |>
  interpolate(hz = 10, max_gap_frames = 20) |>
  filtrate(method = "butterworth", target = "coordinates", cutoff = 1.5) |>
  derivate(derivatives = c("velocity", "acceleration", "angular_velocity"))

trace <- trace_pre |>
  designate(
    method      = "corbett",
    v_threshold = 0.3,
    min_pts     = NULL    # defaults to 5 seconds × Hz = 50 rows
  )

# What segments were found?
table(trace$designation)
```

The example file produces two segments: a long relief period at the
start (the player was indoors with no GPS) and one active bout once they
went outside. In a typical multi-drill session you would see `active_1`,
`relief_1`, `active_2`, `relief_2`, and so on.

### Visualising the segmentation

```{r designate-plot}
trace |>
  ggplot(aes(x = unix_time, y = velocity, colour = designation)) +
  geom_line(linewidth = 0.4) +
  scale_colour_manual(
    values = c(active_1 = "steelblue", relief_1 = "firebrick2"),
    na.value = "grey70"
  ) +
  labs(
    title  = "Velocity coloured by designation",
    x      = "Unix time (s)",
    y      = "Speed (m/s)",
    colour = "Segment"
  ) +
  theme_minimal() +
  theme(panel.grid = element_blank(), legend.position = "top")
```

```{r designate-field, fig.height=5}
# Trajectory coloured by designation — see where each segment occurred spatially
trace |>
  filter(!is.na(f_x)) |>
  ggplot(aes(x = f_x, y = f_y, colour = designation)) +
  geom_path(linewidth = 1.0, alpha = 0.8) +
  coord_fixed() +
  scale_colour_manual(
    values = c(active_1 = "steelblue", relief_1 = "firebrick2"),
    na.value = "grey70"
  ) +
  labs(
    title  = "Trajectory coloured by designation",
    x      = "x (m)", y = "y (m)", colour = NULL
  ) +
  theme_minimal() +
  theme(panel.grid = element_blank(), legend.position = "top")
```

### How the algorithm works

1.  **PELT segmentation.** The velocity signal is scanned for points
    where the mean changes abruptly (SIC penalty prevents
    over-segmentation).
2.  **Active vs relief classification.** Segments with mean velocity
    below `v_threshold` → relief; above → active.
3.  **Short-segment merging.** Segments shorter than `min_pts` rows are
    absorbed into their neighbour.
4.  **Labelling.** Active segments → `active_1`, `active_2`, etc.;
    relief → `relief_1`, `relief_2`, etc.
5.  **Gap propagation.** Rows with NA velocity inherit the nearest
    neighbour's designation.

### Tuning `v_threshold`

| Context                     | Suggested `v_threshold` |
|-----------------------------|-------------------------|
| Field sport (football, AFL) | 0.25–0.5 m/s            |
| Road running / cycling      | 1.0–2.0 m/s             |
| Court sport (netball)       | 0.25–0.5 m/s            |

------------------------------------------------------------------------

## 3. Manual segmentation — `method = 'manual'`

If you have a timing sheet or video annotation with known start/end
times, import those labels directly:

```{r manual, eval=FALSE}
trace <- trace_pre |>
  designate(
    method = "manual",
    file   = "~/data/session_labels.csv"
  )
```

The CSV must have three columns: `designation`, `unix_start`,
`unix_end`.

``` csv
designation,unix_start,unix_end
warm_up,1521942600,1521942900
drill_1,1521942960,1521943320
relief_1,1521943320,1521943440
drill_2,1521943440,1521943800
```

For CSVs with different column names, use `cols = 'guess'` or an
explicit mapping:

```{r manual-map, eval=FALSE}
trace <- trace_pre |>
  designate(
    method = "manual",
    file   = "~/data/annotations.csv",
    cols   = c(
      designation = "label",
      unix_start  = "onset_s",
      unix_end    = "offset_s"
    )
  )
```

------------------------------------------------------------------------

## 5. Exporting automatic designations

Export the auto-generated intervals for review and correction:

```{r export}
intervals <- export_designations(trace)
intervals
```

```{r export-write, eval=FALSE}
# Write to CSV for editing in Excel
export_designations(trace, path = "~/outputs/session_designations.csv")
```

The exported CSV has exactly the format that `method = 'manual'`
expects. A useful review loop:

1.  Run automatic segmentation.
2.  Export intervals to CSV.
3.  Adjust misclassified boundaries in Excel.
4.  Re-run with `method = 'manual'` using the corrected file.

------------------------------------------------------------------------

## 6. Reading the quality log

```{r quality}
quality_report(trace)
```

For `method = 'corbett'`, the quality log records the number of
changepoints detected, their indices, segment lengths, and the
`v_threshold` and `min_pts` settings. For `method = 'manual'`, it
records the source file path and column mapping used.

------------------------------------------------------------------------

## Quick reference

| Argument | Default | Purpose |
|----|----|----|
| `method` | `"corbett"` | `"corbett"` (automatic) or `"manual"` (CSV import) |
| `v_threshold` | `0.25` | m/s — mean below this = relief segment |
| `min_pts` | `5 × hz` | Minimum rows per segment; tiny segments are merged |
| `file` | `NULL` | Path to manual-label CSV (required for `"manual"`) |
| `cols` | `NULL` | Column mapping: `NULL`, `"guess"`, or named vector |
