---
title: "Elaborate: Adding Custom Metrics to a Motion Trace"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Elaborate: Adding Custom Metrics to a Motion Trace}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  warning = FALSE,
  message = FALSE
)
```

```{r pipeline-setup, include=FALSE}
library(motiongrammar)
library(ggplot2)
library(dplyr)

# Build through designate — we demonstrate elaborate() here
trace_pre <- initiate(
  session = system.file("extdata", "example_gps_file_short.csv", package = "motiongrammar"),
  source = 'guess_csv'
) |>
  coordinate(from = "latlong", to = "xyz", crs = 3857, norm = TRUE,
             drop_outliers = TRUE) |>
  interpolate(hz = 10, max_gap_frames = 20) |>
  filtrate(method = "butterworth", target = "coordinates", cutoff = 1.5) |>
  derivate(derivatives = c("velocity", "acceleration", "angular_velocity")) |>
  designate(method = "corbett", v_threshold = 0.3)
```

`elaborate()` is the *motiongrammar* safe version of `dplyr::mutate()`.
It adds or modifies columns while preserving the `motion_trace` class
and all attached attributes (`metadata`, `quality`). To me, this is the
biggest boon of *motiongrammar*, because it's where your creativity can
run free and where you can implement second by second analysis with
ease. For example, you can use elaborate to calculate peak demands (a la
Delaney et al)., or to create your own case_when logic, or even to try
calculating something that you've read about but is not yet supported by
your provider.

------------------------------------------------------------------------

## 1. Why use this over a mutate?

`motion_trace` objects carry quality and meta attributes for
reproducibility. Standard `dplyr::mutate()` silently drops them:

```{r problem, eval=FALSE}
# This loses metadata and quality:
trace <- trace |> 
  dplyr::mutate(curved_load = velocity * abs(angular_velocity))

# This preserves them:
trace <- trace |> 
  elaborate(curved_load = velocity * abs(angular_velocity))
```

Use `elaborate()` any time you want to add a column to a `motion_trace`.

------------------------------------------------------------------------

## 2. Basic usage

Below, I made up an example of "curved load". This is not a real,
validated metric. It just exists to show how you might use your
imagination to create new series of data. If you wanted to have summary
metrics, you'd use quantitate. Elaborate exists to add something to
every data point (ie., create a new time-series).

In the example below, we create a new metric called curve load, we add a
new column that expresses velocity as km/hr and add some logic for an
arbitrary hsr (high speed running) cut off.

```{r basic}
library(motiongrammar)
library(ggplot2)
library(dplyr)

trace_pre <- initiate(
  session = system.file("extdata", "example_gps_file_short.csv", package = "motiongrammar"),
  source = 'guess_csv'
) |>
  coordinate(from = "latlong", to = "xyz", crs = 3857, norm = TRUE,
             drop_outliers = TRUE) |>
  interpolate(hz = 10, max_gap_frames = 20) |>
  filtrate(method = "butterworth", target = "coordinates", cutoff = 1.5) |>
  derivate(derivatives = c("velocity", "acceleration", "angular_velocity")) |>
  designate(method = "corbett", v_threshold = 0.3)

trace <- trace_pre |>
  elaborate(
    curved_load = velocity * abs(angular_velocity),
    velocity_kmh = velocity * 3.6,
    is_hsr = velocity >= 5.5
  )

summary(trace$curved_load)
```

Later expressions can reference columns created earlier in the same
call:

```{r sequential}
trace <- trace_pre |>
  elaborate(
    velocity_kmh = velocity * 3.6,
    is_sprint = velocity_kmh > 25
  )
```

------------------------------------------------------------------------

## 3. Visualising custom metrics

### Curved load on the field trajectory

Here's an example of how we could visualise our new time-series. It
actually looks relatively good, by chance.

```{r curved-load-field-basic, fig.height=5}
trace <- trace_pre |>
  elaborate(curved_load = velocity * abs(angular_velocity))

trace |>
  filter(!is.na(f_x), curved_load < 500) |>
  ggplot(aes(x = f_x, y = f_y, colour = curved_load)) +
  geom_path(linewidth = 1.3) +
  coord_fixed() +
  scale_colour_viridis_c(option = "magma", begin = 0.1, end = 0.95) +
  labs(
    title = "Curved load — velocity × |angular velocity|",
    subtitle = "Bright = high combined velocity and turning",
    x = "x (m)", y = "y (m)", colour = "Curved\nload"
  ) +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

### Curved load time-series

If we wanted to plot curved load as a time-series, we could do the
following

```{r curved-load-ts}
trace |>
  filter(!is.na(f_x)) |>
  ggplot(aes(x = unix_time, y = curved_load)) +
  geom_line(linewidth = 0.3, colour = "black") +
  labs(
    title = "Curved load over time",
    x = "Unix time (s)",
    y = "Curved load (m/s × °/s)"
  ) +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

------------------------------------------------------------------------

## 4. Grouped calculations with `by`

The `by` argument groups the data before mutating and ungroups it after.
This is the mechanism for **within-designation** calculations. If you're
coming from a dplyr background, this is like group_by, but is an
argument in elaborate itself.

### Within-designation normalisation

In our example below, we calculate within the designation. So the max
velocity it finds is within each designation, and not across the entire
session.

```{r by-designation}
trace <- trace_pre |>
  elaborate(
    vel_pct_peak = velocity / max(velocity, na.rm = TRUE) * 100,
    by = "designation"
  )

```

### Time since designation

Or if we want to see how long it's been since the designation began....

```{r by-time}
trace <- trace_pre |>
  elaborate(
    time_in_designation = unix_time - min(unix_time, na.rm = TRUE),
    by = "designation"
  )
```

------------------------------------------------------------------------

## 5. Logging expressions — `log_expr`

Set `log_expr = TRUE` to store the formula text in the quality log. This
is useful for methods sections of publications where you want the full
computational chain embedded in the data object:

```{r log-expr}
trace <- trace_pre |>
  elaborate(
    curved_load = velocity * abs(angular_velocity),
    log_expr = TRUE
  )

# Retrieve the expression from the quality log
attr(trace, "quality")$elaborate[[1]]$expressions
```

------------------------------------------------------------------------

## 6. Case study — Varley Peak Demands

Proposed by Varley and later popularised by Delaney, "peak demand" in
team-sports is sometimes calculated by computing a simple moving average
over an acceleration time-series (and then later summarising it in
quantitate.) I've included an example of how you can use elaborate to
wrap rolling means from zoo to achieve this.

```{r sma-curved-load}
library(zoo)

trace <- trace_pre |>
  elaborate(
    accel_sma = zoo::rollmean(acceleration, k = 10, fill = NA, align = "center"),
  )
```

------------------------------------------------------------------------

## 7. Reading the quality log

```{r quality}
quality_report(trace)
```

The elaborate section records every call: timestamp, columns added,
columns modified, the `by` grouping used, and expression text if
`log_expr = TRUE`.

------------------------------------------------------------------------

## Quick reference

| Argument   | Default | Purpose                                               |
|-------------|-------------|----------------------------------------------|
| `...`      | —       | Column expressions — same syntax as `dplyr::mutate()` |
| `by`       | `NULL`  | Group by these columns before mutating, ungroup after |
| `log_expr` | `FALSE` | Store expression text in the quality log              |

------------------------------------------------------------------------

> **Developer note — preserving the motion_trace structure**
>
> A `motion_trace` is a specialised wrapper class: a tibble with
> `metadata` and `quality` attributes attached. Base R operators that
> build a new data frame from the ground up — `subset()`, `cbind()`,
> `rbind()`, and bracket sub-setting (`df[, 1:3]`) — discard custom
> attributes, silently producing a plain tibble with the class
> declaration and quality log both gone. Calling `dplyr::mutate()`
> directly has the same effect, which is exactly why `elaborate()` exists.
>
> **Always use `elaborate()`** for data transformations and variable
> creation. It calls `dplyr::mutate()` internally but then restores the
> class and all attributes before returning.
>
> **Row-filtering, sorting, and column-selecting** via the standard dplyr
> verbs (`filter()`, `select()`, `arrange()`, `slice()`) are safe:
> *motiongrammar* registers a `vctrs::vec_restore` method that
> reattaches `metadata` and `quality` automatically. If you suspect the
> class has been stripped at any point, check with
> `inherits(trace, "motion_trace")`.
