---
title: "Elaborate: Adding Custom Metrics to a Motion Trace"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Elaborate: Adding Custom Metrics to a Motion Trace}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4,
  warning  = FALSE,
  message  = FALSE
)
```

```{r pipeline-setup, include=FALSE}
library(motiongrammar)
library(ggplot2)
library(dplyr)

# Build through designate — we demonstrate elaborate() here
trace_pre <- initiate(
  session = system.file("extdata", "example_gps_file_short.csv", package = "motiongrammar"),
  source  = 'guess_csv'
) |>
  coordinate(from = "latlong", to = "xyz", crs = 3857, norm = TRUE,
             drop_outliers = TRUE) |>
  interpolate(hz = 10, max_gap_frames = 20) |>
  filtrate(method = "butterworth", target = "coordinates", cutoff = 1.5) |>
  derivate(derivatives = c("velocity", "acceleration", "angular_velocity")) |>
  designate(method = "corbett", v_threshold = 0.3)
```

`elaborate()` is the *motiongrammar* safe version of `dplyr::mutate()`.
It adds or modifies columns while preserving the `motion_trace` class
and all attached attributes (`metadata`, `quality`).

------------------------------------------------------------------------

## 1. The problem it solves

`motion_trace` objects carry quality and meta attributes for
reproducibility. Standard `dplyr::mutate()` silently drops them:

```{r problem, eval=FALSE}
# This loses metadata and quality:
trace <- trace |> dplyr::mutate(curved_load = velocity * abs(angular_velocity))

# This preserves them:
trace <- trace |> elaborate(curved_load = velocity * abs(angular_velocity))
```

Use `elaborate()` any time you want to add a column to a `motion_trace`.

------------------------------------------------------------------------

## 2. Basic usage

Below, I made up an example of "curved load". This is not a real,
validated metric. It just exists to show how you might use your
imagination to create new series of data. If you wanted to have summary
metrics, you'd use quantitate. Elaborate exists to add something to
every data point (ie., create a new time-series)

```{r basic}
library(motiongrammar)
library(ggplot2)
library(dplyr)

trace_pre <- initiate(
  session = system.file("extdata", "example_gps_file_short.csv", package = "motiongrammar"),
  source  = 'guess_csv'
) |>
  coordinate(from = "latlong", to = "xyz", crs = 3857, norm = TRUE,
             drop_outliers = TRUE) |>
  interpolate(hz = 10, max_gap_frames = 20) |>
  filtrate(method = "butterworth", target = "coordinates", cutoff = 1.5) |>
  derivate(derivatives = c("velocity", "acceleration", "angular_velocity")) |>
  designate(method = "corbett", v_threshold = 0.3)

trace <- trace_pre |>
  elaborate(
    curved_load  = velocity * abs(angular_velocity),
    velocity_kmh    = velocity * 3.6,
    is_hsr       = velocity >= 5.5
  )

summary(trace$curved_load)
```

Later expressions can reference columns created earlier in the same
call:

```{r sequential}
trace <- trace_pre |>
  elaborate(
    velocity_kmh  = velocity * 3.6,
    is_sprint  = velocity_kmh > 25
  )
```

------------------------------------------------------------------------

## 3. Visualising custom metrics

### Curved load on the field trajectory

```{r curved-load-field-basic, fig.height=5}
trace <- trace_pre |>
  elaborate(curved_load = velocity * abs(angular_velocity))

trace |>
  filter(!is.na(f_x), curved_load < 500) |>
  ggplot(aes(x = f_x, y = f_y, colour = curved_load)) +
  geom_path(linewidth = 1.3) +
  coord_fixed() +
  scale_colour_viridis_c(option = "magma", begin = 0.1, end = 0.95) +
  labs(
    title    = "Curved load — velocity × |angular velocity|",
    subtitle = "Bright = high combined velocity and turning",
    x = "x (m)", y = "y (m)", colour = "Curved\nload"
  ) +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

### Curved load time-series

```{r curved-load-ts}
trace |>
  filter(!is.na(f_x)) |>
  ggplot(aes(x = unix_time, y = curved_load)) +
  geom_line(linewidth = 0.3, colour = "black") +
  labs(
    title = "Curved load over time",
    x     = "Unix time (s)",
    y     = "Curved load (m/s × °/s)"
  ) +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

------------------------------------------------------------------------

## 4. Grouped calculations with `by`

The `by` argument groups the data before mutating and ungroups it after.
This is the mechanism for **within-designation** calculations.

### Within-designation normalisation

```{r by-designation}
trace <- trace_pre |>
  elaborate(
    vel_pct_peak = velocity / max(velocity, na.rm = TRUE) * 100,
    by = "designation"
  )

```

### Time since designation

```{r by-time}
trace <- trace_pre |>
  elaborate(
    time_in_designation = unix_time - min(unix_time, na.rm = TRUE),
    by = "designation"
  )
```

------------------------------------------------------------------------

## 5. Logging expressions — `log_expr`

Set `log_expr = TRUE` to store the formula text in the quality log —
useful for methods sections of publications where you want the full
computational chain embedded in the data object:

```{r log-expr}
trace <- trace_pre |>
  elaborate(
    curved_load = velocity * abs(angular_velocity),
    log_expr    = TRUE
  )

# Retrieve the expression from the quality log
attr(trace, "quality")$elaborate[[1]]$expressions
```

------------------------------------------------------------------------

## 6. Case study — Varley Peak Demands

Proposed by Varley and later adapted by Delaney, "peak demand" in
team-sports is sometimes calculated by computing a simple moving average
over an acceleration time-series (and then later summarising it in
quantitate.) I've included an example of how you can use elaborate to
wrap rolling means from zoo to achieve this.

```{r sma-curved-load}
library(zoo)

trace <- trace_pre |>
  elaborate(
    accel_sma = zoo::rollmean(acceleration, k = 10, fill = NA, align = "center"),
  )
```

------------------------------------------------------------------------

## 7. Reading the quality log

```{r quality}
quality_report(trace)
```

The elaborate section records every call: timestamp, columns added,
columns modified, the `by` grouping used, and expression text if
`log_expr = TRUE`.

------------------------------------------------------------------------

## Quick reference

| Argument   | Default | Purpose                                               |
|------------|---------|-------------------------------------------------------|
| `...`      | —       | Column expressions — same syntax as `dplyr::mutate()` |
| `by`       | `NULL`  | Group by these columns before mutating, ungroup after |
| `log_expr` | `FALSE` | Store expression text in the quality log              |
