---
title: "Elaborate: Adding Custom Metrics to a Motion Trace"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Elaborate: Adding Custom Metrics to a Motion Trace}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4,
  warning  = FALSE,
  message  = FALSE
)
```

```{r pipeline-setup, include=FALSE}
devtools::load_all()
library(ggplot2)
library(dplyr)

# Build through designate — we demonstrate elaborate() here
trace_pre <- initiate(
  session = '~/motiongrammar/data/example_gps_file_short.csv',
  source  = 'catapult_replay'
) |>
  coordinate(from = "latlong", to = "xyz", crs = 3857, norm = TRUE,
             drop_outliers = TRUE) |>
  interpolate(hz = 10, max_gap_frames = 20) |>
  filtrate(method = "butterworth", target = "coordinates", cutoff = 1.5) |>
  derivate(derivatives = c("velocity", "acceleration", "angular_velocity")) |>
  designate(method = "corbett", v_threshold = 0.3)
```

`elaborate()` is the *motiongrammar*-safe version of `dplyr::mutate()`.
It adds or modifies columns while preserving the `motion_trace` class
and all attached attributes (`metadata`, `quality`).

------------------------------------------------------------------------

## 1. The problem it solves

`motion_trace` objects carry two attributes alongside the data. Standard
`dplyr::mutate()` silently drops them:

```{r problem, eval=FALSE}
# This loses metadata and quality:
trace <- trace |> dplyr::mutate(curved_load = velocity * abs(angular_velocity))

# This preserves them:
trace <- trace |> elaborate(curved_load = velocity * abs(angular_velocity))
```

Use `elaborate()` any time you want to add a column to a `motion_trace`.

------------------------------------------------------------------------

## 2. Basic usage

```{r basic}
library(motiongrammar)
library(ggplot2)
library(dplyr)

trace_pre <- initiate(
  session = '~/motiongrammar/data/example_gps_file_short.csv',
  source  = 'catapult_replay'
) |>
  coordinate(from = "latlong", to = "xyz", crs = 3857, norm = TRUE,
             drop_outliers = TRUE) |>
  interpolate(hz = 10, max_gap_frames = 20) |>
  filtrate(method = "butterworth", target = "coordinates", cutoff = 1.5) |>
  derivate(derivatives = c("velocity", "acceleration", "angular_velocity")) |>
  designate(method = "corbett", v_threshold = 0.3)

trace <- trace_pre |>
  elaborate(
    curved_load  = velocity * abs(angular_velocity),
    speed_kmh    = velocity * 3.6,
    is_hsr       = velocity >= 5.5
  )

summary(trace$curved_load)
```

Later expressions can reference columns created earlier in the same
call:

```{r sequential}
trace <- trace_pre |>
  elaborate(
    speed_kmh  = velocity * 3.6,
    is_sprint  = speed_kmh > 25
  )
```

------------------------------------------------------------------------

## 3. Visualising custom metrics

### Curved load on the field trajectory

```{r curved-load-field, fig.height=5}
trace <- trace_pre |>
  elaborate(curved_load = velocity * abs(angular_velocity))

trace |>
  filter(!is.na(f_x), curved_load < 500) |>
  ggplot(aes(x = f_x, y = f_y, colour = curved_load)) +
  geom_path(linewidth = 1.3) +
  coord_fixed() +
  scale_colour_viridis_c(option = "magma", begin = 0.1, end = 0.95) +
  labs(
    title    = "Curved load — velocity × |angular velocity|",
    subtitle = "Bright = high combined speed and turning",
    x = "x (m)", y = "y (m)", colour = "Curved\nload"
  ) +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

### Curved load time-series

```{r curved-load-ts}
trace |>
  filter(!is.na(f_x)) |>
  ggplot(aes(x = unix_time, y = curved_load)) +
  geom_line(linewidth = 0.3, colour = "darkorchid") +
  labs(
    title = "Curved load over time",
    x     = "Unix time (s)",
    y     = "Curved load (m/s × °/s)"
  ) +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

------------------------------------------------------------------------

## 4. Grouped calculations with `by`

The `by` argument groups the data before mutating and ungroups it after.
This is the mechanism for **within-segment** calculations.

### Within-segment normalisation

```{r by-designation}
trace <- trace_pre |>
  elaborate(
    vel_pct_peak = velocity / max(velocity, na.rm = TRUE) * 100,
    by = "designation"
  )

# Each designation segment now has its own peak as 100%
trace |>
  filter(!is.na(f_x)) |>
  group_by(designation) |>
  summarise(max_pct = max(vel_pct_peak, na.rm = TRUE),
            mean_pct = round(mean(vel_pct_peak, na.rm = TRUE), 1))
```

### Time since segment start

```{r by-time}
trace <- trace_pre |>
  elaborate(
    time_in_segment = unix_time - min(unix_time, na.rm = TRUE),
    by = "designation"
  )
```

------------------------------------------------------------------------

## 5. Logging expressions — `log_expr`

Set `log_expr = TRUE` to store the formula text in the quality log —
useful for methods sections of publications where you want the full
computational chain embedded in the data object:

```{r log-expr}
trace <- trace_pre |>
  elaborate(
    curved_load = velocity * abs(angular_velocity),
    log_expr    = TRUE
  )

# Retrieve the expression from the quality log
attr(trace, "quality")$elaborate[[1]]$expressions
```

------------------------------------------------------------------------

## 6. Practical examples

```{r examples}
trace <- trace_pre |>
  elaborate(
    # Intensity flags
    is_hsr    = velocity >= 5.5,
    is_sprint = velocity >= 7.0,
    # Composite metrics
    curved_load  = velocity * abs(angular_velocity),
    power_proxy  = velocity * pmax(acceleration, 0),   # positive work only
    # Unit conversion
    speed_kmh = velocity * 3.6
  )

# Proportion of active period in HSR
trace |>
  filter(designation == "active_1") |>
  summarise(
    pct_hsr    = round(mean(is_hsr,    na.rm = TRUE) * 100, 1),
    pct_sprint = round(mean(is_sprint, na.rm = TRUE) * 100, 1)
  )
```

------------------------------------------------------------------------

## 7. Reading the quality log

```{r quality}
quality_report(trace)
```

The elaborate section records every call: timestamp, columns added,
columns modified, the `by` grouping used, and expression text if
`log_expr = TRUE`.

------------------------------------------------------------------------

## Quick reference

| Argument   | Default | Purpose                                                         |
|------------|---------|-----------------------------------------------------------------|
| `...`      | —       | Column expressions — same syntax as `dplyr::mutate()`          |
| `by`       | `NULL`  | Group by these columns before mutating, ungroup after          |
| `log_expr` | `FALSE` | Store expression text in the quality log                       |
