---
title: "Filtrate: Removing Noise from Tracking Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Filtrate: Removing Noise from Tracking Data}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4,
  warning  = FALSE,
  message  = FALSE
)
```

```{r pipeline-setup, include=FALSE}
library(motiongrammar)
library(ggplot2)
library(dplyr)
library(tidyr)

# Build trace up to interpolation — we demonstrate filtrate() here
trace_pre <- initiate(
  session = system.file("extdata", "example_gps_file_short.csv", package = "motiongrammar"),
  source  = 'guess_csv'
) |>
  coordinate(from = "latlong", to = "xyz", crs = 3857, norm = TRUE,
             drop_outliers = TRUE) |>
  interpolate(hz = 10, max_gap_frames = 20)
```

GPS and optical tracking data contain measurement noise. Noise presents
as random errors in position that have nothing to do with how the
athlete actually moved. `filtrate()` attenuates (reduces) that noise.
But to use it well, you need to understand *what noise is*, *what a
filter does to it*, and *how to choose the right settings*.

Let me start by saying, that filtering and frequency domain analysis is
a pressure point for most students studying sports science (it certainly
was for me!) It will likely take time for these concepts to sink in, and
for your first few attempts doing this, it's normal to feel like you're
doing some magic without knowing exactly how it works.

------------------------------------------------------------------------

## 1. Introducing Noise

GPS coordinates are not perfect. They have error. This error usually
appears as rapid "jitters" around a players smoother trajectory. Often,
we don't even notice noise until we calculate velocity and acceleration.
If a position error of 0.5 m occurs in one frame (0.1 s at 10 Hz), the
implied velocity error is 0.5 / 0.1 = **5 m/s**. That is larger than
walking pace, introduced from sensor noise. Filtering removes the noise
before differentiation, so the derivatives you compute reflect actual
movement rather than sensor error.

------------------------------------------------------------------------

## 2. The Frequency Domain

To understand filtering, we need to think in terms of frequencies.
*Mathematician Joseph Fourier* proposed that we can think of any signal,
as a bunch of different waves oscillating at different frequencies. So
let's walk through our example file, and see how different waves allow
us to describe motion data...

A GPS position time-series contains:

-   **Low-frequency content (0–2 Hz):** the athlete's actual movement:
    running, turning, decelerating. These patterns change slowly.
-   **High-frequency content (2–5 Hz and above):** measurement noise.
    This changes randomly from frame to frame.

In sports science, we usually use a *low pass* filter. Here we allow the
***low values*** to pass through unchecked. Meanwhile, we *attenuate
(*or reduce) the noise associated with the jitters I mentioned earlier.

### What is a Hertz?

One Hertz (Hz) means one complete cycle per second. A runner who
accelerates and decelerates once every two seconds produces a pattern at
0.5 Hz. Random GPS jitter that flips direction every 0.1 s operates at
10 Hz. The filter separates these two.

### The Nyquist frequency

At a sample rate of 10 Hz, the fastest variation you can measure
oscillates at 5 Hz (changing every 0.1 s). This upper limit is called
the **Nyquist frequency**, which is always exactly half the sample rate.
You cannot specify a cutoff at or above Nyquist and *motiongrammar* will
throw an error if you try as a guardrail.

------------------------------------------------------------------------

## 3. Key filter concepts

### Cutoff frequency

The **cutoff frequency** defines the boundary between what the filter
passes and what it attenuates. A cutoff of 1.5 Hz means: "let through
everything below 1.5 Hz. Reduce everything above it."

### Roll-off

**Roll-off** describes how steeply the filter attenuates signals above
the cutoff. It is measured in **decibels per octave** (dB per doubling
of frequency).

If the cutoff is 2 Hz. A signal at 2 Hz passes at full strength. A
signal at 4 Hz (one octave above cutoff) is attenuated by the roll-off
amount. For a 2nd-order Butterworth filter the roll-off is approximately
12 dB/octave — meaning the 4 Hz signal exits at about one-quarter of its
original power. At 8 Hz (two octaves up) it is one-sixteenth.

**Higher filter order → steeper roll-off → sharper cutoff**, but with
increasing risk of ringing artefacts near the cutoff frequency. Second
order is the biomechanics standard for this reason.

### Attenuation

**Attenuation** is how much a filter reduces the amplitude of a signal
at a given frequency, expressed in decibels:

```         
Attenuation (dB) = 20 × log10(output_amplitude / input_amplitude)
```

| dB | What it means |
|------------------------------------|------------------------------------|
| 0 dB | No change — signal passes unchanged |
| −3 dB | Amplitude reduced to 70.7% — this is the standard definition of the cutoff frequency for Butterworth |
| −20 dB | Amplitude reduced to 10% |
| −40 dB | Amplitude reduced to 1% |

When someone says "the cutoff is 1.5 Hz", they mean the signal at 1.5 Hz
is attenuated by 3 dB, or is reduced to \~70% of its original amplitude.
Frequencies well above the cutoff are attenuated far more.

### Phase shift

A causal filter (one that only looks backward in time, like EMA)
introduces a **phase lag** — the filtered output trails the raw input in
time. This distorts the timing of acceleration peaks and turning events.
This is why most of the time you will hear people avoiding them, and
using a *Butterworth filter*, even though it is regerettably a lot more
difficult to understand.

The Butterworth filter in *motiongrammar* is applied **zero-phase** via
`signal::filtfilt()` — forward and backward passes cancel each other's
phase shift, leaving the filtered signal perfectly time-aligned with the
raw data.

------------------------------------------------------------------------

## 4. Finding the right cutoff using `filtrate_cutoff()`

### 4.1 Residual analysis (Winter, 2009)

This method sweeps a range of cutoff frequencies, applies the filter at
each one, and computes the RMS of the residual (raw − filtered). When
the cutoff is low, too much real signal is removed and the residual is
large. As you raise it, the residual drops until only noise remains —
the residual flattens. The **elbow of that curve** is the optimal
cutoff.

```{r load-for-cutoff, include=FALSE}
library(motiongrammar)
library(ggplot2)
library(dplyr)
library(tidyr)

trace_pre <- initiate(
  session = system.file("extdata", "example_gps_file_short.csv", package = "motiongrammar"),
  source  = 'guess_csv'
) |>
  coordinate(from = "latlong", to = "xyz", crs = 3857, norm = TRUE,
             drop_outliers = TRUE) |>
  interpolate(hz = 10, max_gap_frames = 20)
```

```{r cutoff-residual, fig.height=4}
filtrate_cutoff(
  trace_pre,
  method       = "butterworth",
  target       = "coordinates",
  axis         = "x",
  plot_type    = "residual",
  cutoff_range = c(0.1, 4),
  n_steps      = 60
)
```

The dashed red line marks the suggested cutoff (point of maximum
curvature of the smoothed residual curve). Use it as a starting point,
not a hard rule. In reality, you only have to work out a cut off once.
In every team sport setting I have worked in, we *designed a filter* and
identified a cut off that works well across the sample we had, rather
than designing and applying a brand new filter to every GPS file.

### 4.2 Power Spectral Density (PSD)

The PSD shows how signal power is distributed across frequencies. Real
movement energy clusters at low frequencies; noise is relatively flat at
higher frequencies. Use this to confirm where the signal ends.

```{r cutoff-psd, fig.height=4}
filtrate_cutoff(
  trace_pre,
  method    = "butterworth",
  target    = "coordinates",
  axis      = "x",
  plot_type = "psd"
)
```

### 4.3 Side-by-side

```{r cutoff-both, fig.height=5}
filtrate_cutoff(
  trace_pre,
  method       = "butterworth",
  target       = "coordinates",
  axis         = "x",
  plot_type    = "both",
  cutoff_range = c(0.1, 4),
  n_steps      = 60
)
```

A cutoff of **1–2 Hz** is typical for GPS position data at 10 Hz. If the
residual plot declines like a straight line (ie., no bend in the elbow),
the data was likely already filtered by the device before export. But
it's always good to check!

------------------------------------------------------------------------

## 5. Should we filter coordinates or derivatives?

This is a common debate in biomechanics, that I am not fully qualified
to contribute to. Therefore, I support both options in *motiongrammar*,
with a slight preference to filtering coordinates rather than
derivatives. My reasoning for this is because in *motiongrammar* we deal
with velocity, acceleration and angular velocity– all of which operate
on different units. It is smoother to work out a filter on xyz for all
of these, versus running different filters on derivatives. But you do
you!

**Filter coordinates first, then differentiate** (Winter, 2009): the
position signal is the primary measurement; smoothing it once propagates
consistently to all derivatives.

**Differentiate first, then filter derivatives directly**: proponents
argue this isolates noise in the derivative domain where it is easier to
see.

*motiongrammar* supports both via the `target` argument. The recommended
default is to filter coordinates. Whatever you choose, **report it in
any publication**: the method, order, and cutoff frequency.

------------------------------------------------------------------------

## 6. Applying the filter

```{r filtrate}
trace <- trace_pre |>
  filtrate(
    method = "butterworth",
    target = "coordinates",
    n      = 2,       # 2nd order — standard in biomechanics
    cutoff = 1.5,     # Hz — chosen from filtrate_cutoff() above
    type   = "low"
  )
```

### Raw vs filtered comparison

```{r compare-plot}
# Pull out a representative 60-second window with GPS data
window <- trace |>
  filter(!is.na(f_x)) |>
  slice(1:600)

window |>
  ggplot(aes(x = unix_time)) +
  geom_line(aes(y = x,   colour = "Raw"),      linewidth = 0.5) +
  geom_line(aes(y = f_x, colour = "Filtered"), linewidth = 0.5) +
  scale_colour_manual(values = c(Raw = "firebrick2", Filtered = "steelblue")) +
  labs(
    title  = "x coordinate — raw vs Butterworth filtered (1.5 Hz, 2nd order)",
    x      = "Unix time (s)",
    y      = "x (m)",
    colour = NULL
  ) +
  theme_minimal() +
  theme(panel.grid = element_blank(), legend.position = "top")
```

```{r compare-traj}
# Filtered trajectory is smoother — compare the path shapes
bind_rows(
  trace |> filter(!is.na(x))   |> select(x, y)   |> mutate(type = "Raw"),
  trace |> filter(!is.na(f_x)) |> select(x = f_x, y = f_y) |> mutate(type = "Filtered")
) |>
  ggplot(aes(x = x, y = y, colour = type)) +
  geom_path(alpha = 0.6, linewidth = 0.5) +
  coord_fixed() +
  scale_colour_manual(values = c(Raw = "firebrick2", Filtered = "steelblue")) +
  labs(
    title  = "Trajectory — raw vs filtered",
    x      = "x (m)", y = "y (m)", colour = NULL
  ) +
  theme_minimal() +
  theme(panel.grid = element_blank(), legend.position = "top")
```

If the two trajectories are nearly identical, the provider already
filtered the data before export.

------------------------------------------------------------------------

## 7. Available filter methods

### Butterworth (recommended)

Zero-phase IIR filter. Applied forward and backward via
`signal::filtfilt()` so the output is time-aligned with the input.
Standard in biomechanics (Winter, 2009).

**Key parameters:**

| Parameter | Guidance |
|------------------------------------|------------------------------------|
| `n` | Filter order. 2nd order gives 40 dB/decade rolloff with no ringing. |
| `cutoff` | Must be between 0 and Nyquist (sample_rate / 2). |
| `type` | `"low"` for noise removal (almost always). |

### Simple Moving Average (SMA)

Replaces each value with the mean of the surrounding `window` points. At
10 Hz, `window = 10` averages over 1 second. Gentle roll-off —
attenuates noise gradually rather than sharply. Edge rows become `NA`.
This introduces the phase shift we mentioned earlier, but if you do not
care about time alignment (for example you're only calculating aggregate
parameters) then this is a reasonable option.

```{r sma, eval=FALSE}
trace |> filtrate(method = "sma",
                  target = "coordinates",
                  window = 5)
```

### Exponential Moving Average (EMA)

Weights past values exponentially using a dampening ratio. This
introduces phase lag. I have seen this used in real time streaming
because it's relatively fast, but I do not see a clear use case ad-hoc.

```{r ema, eval=FALSE}
trace |> filtrate(method = "ema", 
                  target = "coordinates", 
                  alpha = 0.3)
```

### Savitzky-Golay

Fits a polynomial through a sliding window. Better at preserving peaks
and troughs than SMA because a polynomial can still pass through local
maxima. Requires the `signal` package. I have not seen this used too
much in Sports Science, but some of my post PhD reading suggests it
could be a good option to attenuate signals and reduce some of those
enormous velocity spikes that still seem to pass through with GPS data.
Someone out there – write a paper on this ;).

```{r savgol, eval=FALSE}
trace |> filtrate(method = "savgol", target = "coordinates",
                  window = 11, poly_order = 3)
```

------------------------------------------------------------------------

## 8. The `f_` columns

For each filtered column, `filtrate()` creates a new `f_` prefixed
column. The raw column is always kept.

| Raw        | Filtered     |
|------------|--------------|
| `x`        | `f_x`        |
| `y`        | `f_y`        |
| `velocity` | `f_velocity` |

`derivate()` automatically uses `f_x` / `f_y` when they exist.

------------------------------------------------------------------------

## 9. Targeting specific signals

If you want to target coordinates, or a specific derivative you can
specify that here. For multiple targets, you'll need to use a vector.

```{r target, eval=FALSE}
# Coordinates (default)
trace |> filtrate(method = "butterworth", target = "coordinates", cutoff = 1.5)

# Velocity only (use after derivate())
trace |> filtrate(method = "butterworth", target = "velocity", cutoff = 2.0)

# Both velocity and acceleration
trace |> filtrate(method = "butterworth",
                  target = c("velocity", "acceleration"), cutoff = 2.0)
```

------------------------------------------------------------------------

## 10. Reading the quality log

```{r quality}
quality_report(trace,
               step = 'filtrate.')
```

The filtrate section records every pass: method, parameters, which
source columns were filtered, NAs introduced by edge effects, and
package versions for full reproducibility.

------------------------------------------------------------------------

## Quick reference

| Argument     | Default         | Purpose                                    |
|----------------|----------------|---------------------------------------|
| `method`     | `"butterworth"` | Filter algorithm                           |
| `target`     | `"coordinates"` | Signal(s) to filter                        |
| `n`          | `2`             | Butterworth filter order                   |
| `cutoff`     | `1`             | Butterworth cutoff frequency (Hz)          |
| `type`       | `"low"`         | Butterworth filter type                    |
| `window`     | `5`             | SMA or Savitzky-Golay window size (rows)   |
| `alpha`      | `0.3`           | EMA smoothing factor (0 = heavy, 1 = none) |
| `poly_order` | `2`             | Savitzky-Golay polynomial order            |

