---
title: "Initiate: Loading Data into motiongrammar"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Initiate: Loading Data into motiongrammar}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  warning = FALSE,
  message = FALSE
)
```

We enter the world of *motiongrammar* by initiating our dataset.
`initiate()` reads raw tracking data from any supported source and
returns a **`motion_trace`** object. A **`motion_trace`** is a tidy
tibble with `metadata` and `quality` attributes that travel with data
moving forward. This is our import step.

This vignette covers supported data sources, explains the template
system for reproducible workflows, and closes with batch processing a
folder of GPS files.

------------------------------------------------------------------------

## 1. The motion_trace object

Every call to `initiate()` returns the same structure regardless of
source:

| Component | What it contains |
|----|----|
| **Data** | A tibble with `unix_time` plus coordinate columns (`lat`/`lng`/`altitude` for GPS, or `x`/`y`/`z` for local). Extra columns (velocity, heart rate, etc.) appear after the core columns. |
| **Metadata** | A named list include sample rate (Hz), duration, device info, player/session identifiers. Access with `metadata_report()`. We save memory by not requiring dedicated columns for static metadata. |
| **Quality** | An audit log seeded at ingestion and appended by every subsequent pipeline step. Access with `quality_report()`. |

```{r show-structure, eval=FALSE}
trace <- initiate(session = "player_01.csv", method = "guess")

# Print a concise summary
trace

# Inspect metadata
metadata_report(trace)

# Inspect the quality log (what did initiate detect about this file?)
quality_report(trace)
```

The quality log records things like coordinate completeness, sampling
rate, temporal gaps, duplicate timestamps, and any automatic flags. It
also records every subsequent transformation *you* do on the data. It is
updated automatically, so you never need to write to it yourself.

------------------------------------------------------------------------

## 2. CSV files

CSV is the most common format for GPS and optical tracking exports.
*motiongrammar* offers three modes for reading CSV data, ranging from
fully automatic to fully explicit.

### 2.1 `guess_csv` for exploring data

```{r guess}
trace <- initiate(
  session = "~/data/player_01.csv",
  source = "guess_csv"
)
```

The guess algorithm:

1.  Detects the start of the data block by finding the first row whose
    comma count matches the modal comma count of the file. This strips
    preamble rows automatically.
2.  Searches for timestamp, latitude, longitude, and altitude columns
    using a prioritised list of exact and fuzzy-matched names (e.g.
    `"Timestamp"`, `"UTC"`, `"frame_id"` → `unix_time`).
3.  Warns you if any column could not be resolved.

`source = 'auto'` (the default when you supply a `.csv` path) routes to
`guess_csv`:

```{r auto}
# Equivalent — extension triggers guess_csv automatically
trace <- initiate(session = "~/data/player_01.csv")
```

**When to use:** One-off exploration, checking a new file format for the
first time. Not recommended for production pipelines because the
resolved column mapping can silently change if the file layout changes.
The fuzzy logic is also not fool proof so it is always important to
check!

### 2.2 `manual_csv` — explicit column names

```{r manual-csv}
trace <- initiate(
  session = "~/data/player_01.csv",
  source = "manual_csv",
  col_unix = "Timestamp",
  col_lat = "Latitude",
  col_lng = "Longitude",
  col_altitude = "Elevation"
)
```

You name the columns explicitly; *motiongrammar* does the rest
(timestamp parsing, unit coercion, gap detection). Column matching is
case-insensitive.

Useful additional arguments:

| Argument | Default | Purpose |
|----|----|----|
| `skip` | `0` | Header rows to skip before column names row (eg. catapult used to have some meta in their csv) |
| `max_empty_lines` | `3` | Empty rows that signal end-of-data |
| `comment` | `"#"` | Lines starting with this character are ignored |

**When to use:** When the column names are stable, but you are still
playing around/not yet fully comfortable with the supported template
system (outlined below).

### 2.3 `template_csv` — reproducible production mode

A `csv_template` locks the column mapping into an object that can be
saved to XML, version-controlled, and loaded in any future script
without relying on heuristics.

#### Creating a template manually

```{r csv-template-manual}
my_template <- csv_template(
  coord_system = "gps",
  col_unix = "Timestamp",
  col_lat = "Latitude",
  col_lng = "Longitude",
  col_altitude = "Elevation",
  skip = 0L
)

# Print to verify the mapping
my_template
```

You can also carry extra columns through. So for example, if you wanted
to preserve something like heart rate, metabolic power or Player Load.

```{r csv-template-extra}
my_template <- csv_template(
  coord_system = "gps",
  col_unix = "Timestamp",
  col_lat = "Latitude",
  col_lng = "Longitude",
  col_altitude = "Elevation",
  col_velocity = "Speed",          # imported as `velocity`
  col_acceleration = "Accel",         # imported as `acceleration`
  col_extra = c(
    player_load = "PlayerLoad",      # renamed to `player_load`
    heart_rate = "HR_bpm"           # renamed to `heart_rate`
  )
)
```

#### Generating a template automatically from a file

If you already have a file that `guess_csv` reads correctly, capture the
resolved schema in one step:

```{r guess-csv-template}
# Inspect and save in one call
my_template <- guess_csv_template(
  path = "~/data/player_01.csv",
  save_path = "~/templates/catapult_gps.xml"
)

# The saved XML can now be committed to version control
```

Check the returned template. If any column was mis-detected, tweak it
with `csv_template()` and re-save. You can also just open the xml in a
text editor and manually change the inputs.

#### Saving and loading templates

```{r save-load-template}
# Save to XML using this function
save_csv_template(my_template,
                  path = "~/templates/catapult_gps.xml")

# Then in a future script, you can load it like this
my_template <- load_csv_template("~/templates/catapult_gps.xml")
```

#### Using the template with `initiate()`

```{r initiate-template}
# Pass the in-memory object
trace <- initiate(session = "~/data/player_01.csv", 
                  template = my_template)

# Or pass the file path to the template directly.
trace <- initiate(
  session = "~/data/player_01.csv",
  template = "~/templates/catapult_gps.xml"
)
```

When a `template` argument is supplied, `source` is automatically set to
`'template_csv'`.

------------------------------------------------------------------------

## 3. GPX files

GPX is the standard exchange format for consumer GPS devices (Garmin,
Wahoo, Polar, Apple Watch, etc.).

```{r gpx}
trace <- initiate(
  session = "~/data/run_2024-09-14.gpx",
  source = "gpx"
)

# Or let 'auto' detect the extension
trace <- initiate(session = "~/data/run_2024-09-14.gpx")
```

Internally *motiongrammar* uses the `sf` package to read the
`track_points` layer. The `ele` (elevation) field is imported as
`altitude`; timestamp is converted to Unix seconds. The `sf` package is
not listed as a hard dependency, but if it isn't installed you'll see a
clear error message with the install command.

**Note:** If your GPX contains multiple tracks (e.g. a file exported
after pausing and restarting), all track points are concatenated in
chronological order. Large temporal gaps between tracks will be flagged
in the quality log.

------------------------------------------------------------------------

## 4. Strava

*motiongrammar* fetches GPS streams directly from the Strava v3 API. You
need a one-time OAuth2 setup to obtain a `~/strava_tokens.json` file
(client ID, client secret, refresh token). After that, token refresh is
handled automatically. I have a separate guide on this, so here we will
just focus on key syntax once you're set up.

```{r strava-basic}
# Pass the Strava activity ID (or the full activity URL)
trace <- initiate(
  session = "12345678901",
  source = "strava"
)

# Full URL also works
trace <- initiate(
  session = "https://www.strava.com/activities/12345678901",
  source = "strava"
)
```

By default, *motiongrammar* requests `time`, `latlng`, and `altitude`
streams. Activity metadata (name, sport type, start date, elapsed time)
is fetched from the activity endpoint and stored in the `metadata`
attribute automatically.

### Customising Strava streams with `api_stream_template`

If you want to pull additional streams (heart rate, power, cadence) or
store your credentials somewhere other than `~/strava_tokens.json`,
build an `api_stream_template`:

```{r strava-template}
strava_my_template <- api_stream_template(
  api = "strava",
  token_path = "~/credentials/strava.json",
  stream_time = "time",
  stream_latlng = "latlng",
  stream_altitude = "altitude",
  stream_extra = c(
    heart_rate = "heartrate",   # rename to heart_rate
    watts = "watts"
  )
)

trace <- initiate(
  session = "12345678901",
  template = strava_my_template
)
```

------------------------------------------------------------------------

## 5. Generic REST API

Any JSON REST API can be connected via `api_stream_template`. The
template captures the endpoint URL, authentication method, and
field-to-column mapping. Once saved, any collaborator can reproduce the
exact same ingestion without knowing the implementation details. My
suggestion here is to reach out to your vendor. They are unlikely to be
familiar with *motiongrammar* directly, but they will be better placed
to answer questions on how you authenticate etc.

### 5.1 Response formats

*motiongrammar* handles three JSON response layouts:

**`"records"`** — an array of row objects:

``` json
[{"time": 1700000000, "lat": 51.5, "lng": -0.1}, ...]
```

**`"columnar"`** — an object of parallel arrays:

``` json
{"time": [1700000000, ...], "lat": [51.5, ...], "lng": [-0.1, ...]}
```

**`"streams"`** — Strava-style, each key is an object with a `"data"`
array:

``` json
{"time": {"data": [1700000000, ...]}, "latlng": {"data": [[51.5, -0.1], ...]}}
```

### 5.2 Authentication options

| `auth_type`        | How it works                               |
|--------------------|--------------------------------------------|
| `"none"`           | No authentication                          |
| `"bearer"`         | `Authorization: Bearer <token>` header     |
| `"api_key_header"` | Custom header (default `X-API-Key: <key>`) |
| `"api_key_query"`  | Query parameter (default `?api_key=<key>`) |

If `auth_value` starts with `$`, it is treated as an environment
variable name — the token is read at runtime rather than stored in the
template.

### 5.3 Example: records-format API with bearer auth

```{r generic-api-records}
my_template <- api_stream_template(
  api = "generic",
  coord_system = "gps",
  url = "https://api.myprovider.io/sessions/{session}/gps",
  auth_type = "bearer",
  auth_value = "$GPS_API_TOKEN",   # read from env var
  response_format = "records",
  stream_time = "timestamp",
  stream_lat = "lat",
  stream_lng = "lng",
  stream_altitude = "alt",
  stream_extra = c(speed = "velocity_ms")
)

trace <- initiate(session = "SESSION_ID_123", template = my_template)
```

### 5.4 Example: columnar-format API nested inside a response wrapper

```{r generic-api-columnar}
my_template <- api_stream_template(
  api = "generic",
  coord_system = "gps",
  url = "https://api.myprovider.io/v2/tracks/{session}",
  auth_type = "api_key_query",
  auth_value = "$PROVIDER_KEY",
  auth_param = "key",
  response_format = "columnar",
  data_path = "result.data",     # navigate to resp$result$data
  stream_time = "ts",
  stream_lat = "latitude",
  stream_lng = "longitude"
)
```

### 5.5 Saving and reusing API templates

```{r api-template-save}
save_api_stream_template(my_template, path = "~/templates/my_provider.xml")

# Load later — or pass the path directly to initiate()
trace <- initiate(
  session = "SESSION_ID_123",
  template = "~/templates/my_provider.xml"
)
```

------------------------------------------------------------------------

## 6. Enriching metadata after ingestion

`initiate()` populates metadata automatically (session name, sampling
rate, duration, source, etc.), but fields like player name, team, and
sport are left blank. You may fill those in with `set_metadata()`:

```{r set-metadata}
trace <- trace |>
  set_metadata(
    player_id = "P42",
    player_name = "Jade Haycraft",
    team = "Senior A",
    sport = "Ice Skating",
    session_type = "Training"
  )

metadata_report(trace)
```

All keys you pass to `set_metadata()` are added (or updated) on the
metadata list. The quality log records that ingestion happened but does
not prevent you from adding metadata at any point in the pipeline.

------------------------------------------------------------------------

## 7. Batch processing — loading a folder of GPS files

A common workflow is to process an entire match or training day where
each player has their own file. I suggest using `purrr::map()` to apply
the full ingestion pipeline over a folder, producing a list of
`motion_trace` objects that can be analysed or exported together. Be
careful here, because memory usage can spiral out of control. If you're
not working on a server or dealing with long files/high sample rates, I
would probably suggest using the flow control of your choice (whether
that's purrr, a loop or oldschool lapply) to run your entire workflow
one game at a time.

### 7.1 Flat list of traces

```{r batch-list}
library(tidyverse)
library(motiongrammar)

# Load a shared template once
my_template <- load_csv_template("~/templates/catapult_gps.xml")

# Collect all CSV files in the session folder
files <- list.files(
  path = "~/data/session_2024-09-14/",
  pattern = "\\.csv$",
  full.names = TRUE
)

# Map over the files — each becomes a motion_trace
traces <- map(files, \(f) {
  initiate(session = f, template = my_template) |>
    set_metadata(
      player_name = tools::file_path_sans_ext(basename(f)),
      session_type = "Match"
    )
})

names(traces) <- tools::file_path_sans_ext(basename(files))
```

------------------------------------------------------------------------

## 8. Converting an in-memory data frame

Sometimes your tracking data is already in R — pulled from a database,
returned by an API wrapper, produced by a simulation, or loaded with
your own `read_csv()` call. In these cases you don't want to write the
data out to a file just to read it back through `initiate()`. Use
`as_motion_trace()` instead.

### 8.1 GPS data (lat / lng)

```{r as-motion-trace-gps}
# Imagine this came from a database query or a bespoke parser
raw_df <- data.frame(
  timestamp = seq(1700000000, by = 0.1, length.out = 1000),
  latitude = 51.5 + cumsum(rnorm(1000, 0, 0.00005)),
  longitude = -0.1 + cumsum(rnorm(1000, 0, 0.00005)),
  altitude = 12
)

trace <- as_motion_trace(
  raw_df,
  coord_system = "gps",
  col_time = "timestamp",
  col_lat = "latitude",
  col_lng = "longitude",
  # Extra metadata passed straight through
  player_name = "Jade Haycraft",
  sport = "Football",
  session_type = "Match"
)

trace
```

The result is a fully formed `motion_trace` with `metadata` and
`quality` populated. From here you continue with the normal pipeline —
`coordinate()`, `interpolate()`, `derivate()`, etc.

### 8.2 Local (x / y) coordinate data

Optical and UWB tracking systems often export positions in metres
relative to a fixed origin (e.g. the centre spot) rather than in
latitude and longitude. Pass `coord_system = "local"`:

```{r as-motion-trace-local}
raw_local <- data.frame(
  t = seq(0, 99.9, by = 0.1),
  pos_x = cumsum(rnorm(1000, 0.05, 0.2)),
  pos_y = cumsum(rnorm(1000, 0.01, 0.2))
)

trace <- as_motion_trace(
  raw_local,
  coord_system = "local",
  col_time = "t",
  col_x = "pos_x",
  col_y = "pos_y",
  player_name = "Alex Burnett",
  sport = "Basketball"
)
```

Because the data is already in Cartesian metres, you can skip
`coordinate()` and proceed directly to `interpolate()` → `derivate()`.

### 8.3 Column auto-detection

If your column names already match common conventions you can omit the
`col_*` arguments entirely:

```{r as-motion-trace-autodetect}
# Columns named 'time', 'lat', 'lng' → detected without hints
trace <- as_motion_trace(df, coord_system = "gps")

# Columns named 'x', 'y' → detected without hints
trace <- as_motion_trace(df, coord_system = "local")
```

`as_motion_trace()` tries these names in priority order:

| Role | Searched names (in order)                              |
|------|--------------------------------------------------------|
| Time | `unix_time`, `time`, `timestamp`, `t`, `unix`, `epoch` |
| Lat  | `lat`, `latitude`                                      |
| Lng  | `lng`, `lon`, `longitude`                              |
| X    | `x`                                                    |
| Y    | `y`                                                    |
| Z    | `z` (optional)                                         |

If a required column cannot be detected, the function stops with a clear
error message that tells you which `col_*` argument to supply.

------------------------------------------------------------------------

## Quick reference

| Source | Function / `source =` | When to use |
|----|----|----|
| CSV (auto-detect columns) | `initiate()` `"guess_csv"` / `"auto"` | Exploration and first look at a new file |
| CSV (explicit columns) | `initiate()` `"manual_csv"` | Stable column names, no saved template yet |
| CSV (locked schema) | `initiate()` `"template_csv"` | Production, reproducible, team pipelines |
| Catapult replay export | `initiate()` `"catapult_replay"` | Catapult-specific preamble format |
| GPX track file | `initiate()` `"gpx"` / `"auto"` | Consumer GPS devices, Garmin, Apple Watch |
| Strava API | `initiate()` `"strava"` | Activity data from the Strava platform |
| Any JSON REST API | `initiate()` `"generic"` | Custom tracking platform or in-house system |
| In-memory data frame | `as_motion_trace()` | Data already in R (database query, simulation, bespoke parser) |

The recommended template workflow for any repeated analysis:

```         
guess_csv_template()   →   inspect & adjust   →   save_csv_template()
                                                         ↓
                               initiate(template = "my_device.xml")
```

This ensures that column parsing is deterministic, documented, and
independent of heuristics — exactly what you need when the analysis will
be reported, peer-reviewed, or run across many sessions.

------------------------------------------------------------------------

> **Developer note — keeping attributes intact**
>
> A `motion_trace` is a standard tibble with two extra attributes
> attached: `metadata` and `quality`. Base R operators that construct a
> new data frame from scratch — `subset()`, `cbind()`, `rbind()`, and
> bracket sub-setting (`df[, 1:3]`) — strip custom attributes, silently
> returning a plain tibble with no quality log and no class declaration.
> Calling `dplyr::mutate()` directly has the same effect.
>
> **To add or modify columns** and keep the attributes intact, always use
> `elaborate()`.
>
> **For row-filtering, column-selecting, sorting, and slicing**, the
> standard dplyr verbs (`filter()`, `select()`, `arrange()`, `slice()`)
> are safe — *motiongrammar* registers a `vctrs::vec_restore` method that
> reattaches `metadata` and `quality` automatically after any dplyr
> operation. If you ever suspect the class has been dropped, check with
> `inherits(trace, "motion_trace")`.
