---
title: "Initiating Motion Traces from APIs"
subtitle: "Strava defaults, extended streams, and generic REST templates"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  eval = FALSE
)
```

# Overview

`initiate()` can read motion data from local files such as CSV or GPX,
but it can also pull data directly from an API. In this vignette we
cover API support in detail.

There are two main pathways:

1.  **Dedicated Strava pathway**: the simplest option for Strava
    activities. Uses `source = "strava"`, fetches activity metadata and
    stream data, and handles token refresh automatically.
2.  **Generic REST pathway**: a flexible template-based pathway for any
    JSON API. Useful whenever an API returns time-series data in a
    predictable structure that can be mapped to the standard
    `motion_trace` columns.

Recommended learning sequence:

1.  Start with Strava defaults.
2.  Add extra Strava streams (heart rate, distance, cadence, power).
3.  Use Strava as a test case for the generic REST machinery.
4.  Adapt the same template idea to other APIs.

# How APIs work

When you connect to a platform like Strava or WHOOP, you are not
downloading a file. You are sending a request to a remote server and
asking it to send your data back. That server is reached through an
**API** (Application Programming Interface).

## Authentication

Most APIs require you to prove who you are before they hand over data.
They do this with **tokens**. Present the right token with your request
and the server returns your data.

There are two main patterns:

**Simple key (API key/bearer token)**\
The platform gives you a single key that stays valid indefinitely. You
copy the key from the platform's settings page, store it somewhere safe,
and the package attaches it to every request automatically.

**OAuth 2.0 (short lived tokens)**\
Consumer platforms like Strava, Oura, Garmin, and WHOOP use a more
secure two-token system. The **access token** (what gets attached to
each request) expires after a few hours. The **refresh token**
(long-lived, obtained when you first authorise the app through the
browser) is used to silently obtain a new access token when the old one
expires. This is what `auth_type = "oauth2_refresh"` handles
automatically.

## Use a template, not inline credentials

**Never paste a token or API key directly into your R script.**

If a credential appears in your code, it is a single `git push` away
from being public. Once a token leaks, anyone can have it. Save yourself
the headache in the future, by keeping your credentials out of your
code.

The motiongrammar pattern keeps credentials completely out of your code:

-   Store credentials in your `.Renviron` file (lines like
    `MY_TOKEN=abc123`). R loads this at startup but it is never
    committed to version control — add it to `.gitignore` if it is in
    your project folder.
-   Reference credentials in templates as `auth_value = "$MY_TOKEN"`.
    The `$` prefix means "read this from the environment at runtime."
    The template file itself is safe to commit because it contains only
    the variable name, not the value.
-   For OAuth2 providers, store tokens in a JSON file (e.g.
    `~/strava_tokens.json`) and add that file to `.gitignore`. The
    package reads and refreshes the file automatically.

Saving templates as XML files and referencing credentials via
environment variables means your scripts and templates can be
version-controlled, shared with collaborators, and checked into GitHub
without exposing anything sensitive.

NB: you also need to ensure your template is not being pushed to github,
This is vital, as it will also leak your credentials. I recommend saving
the file outside of your workspace, or adding \*.xml to your git ignore.

# Prerequisites

```{r}
library(motiongrammar)
```

For the Strava pathway, you need a token file at:

```{r}
path.expand("~/strava_tokens.json")
```

The token file should look like this (replace with your own values):

``` json
{
  "client_id":     "98765",
  "client_secret": "a1b2c3d4e5f6a1b2c3d4e5f6a1b2c3d4e5f6a1b2",
  "refresh_token": "fc8d2e1a...long string...",
  "access_token":  "eyJhbGci...long string...",
  "expires_at":    0
}
```

Where each field comes from:

| Field | What it is | Where to get it |
|---|---|---|
| `client_id` | A number that identifies your registered Strava app | Strava developer portal → My API Application |
| `client_secret` | A secret key tied to your app (not your Strava password) | Same page as client_id |
| `refresh_token` | A long-lived token that lets the package get fresh access tokens | Generated during the initial OAuth authorisation step |
| `access_token` | The short-lived token actually used in API requests | Set to any value — the package replaces it on the first run |
| `expires_at` | Unix timestamp for when the access token expires | Set to `0` to force a refresh on the first run |

The `client_id` and `client_secret` are not your Strava login — they
identify the application you registered on Strava's developer portal.
See Section 6 for a full explanation of the OAuth2 fields.

# 1. Strava: default API initiation

The cleanest Strava usage. Uses the package's dedicated Strava branch
and default stream mapping.

```{r}
run_default <- initiate(
  source = "strava",
  session = "18529469417"
)

run_default
quality_report(run_default)
metadata_report(run_default)
```

You can also pass a full Strava activity URL:

```{r}
run_default <- initiate(
  source = "strava",
  session = "https://www.strava.com/activities/18529469417"
)
```

By default, the Strava pathway requests these streams:

-   `time`
-   `latlng`
-   `altitude`

Output columns:

``` text
unix_time, lat, lng, altitude, is_interpolated
```

The dedicated Strava path also fetches activity metadata first, which
includes the session start time. That allows the package to convert
Strava's relative stream time (seconds elapsed from the activity start)
into real Unix time. The generic REST path does not do this
automatically.

# 2. Strava: adding extra streams

Strava streams can include additional fields:

-   `heartrate`
-   `distance`
-   `velocity_smooth`
-   `cadence`
-   `watts`

To request these, use an `api_stream_template()` with `api = "strava"`.
This still uses the dedicated Strava pathway, but the template tells it
which extra streams to request and how to name them in the output.

```{r}
strava_template <- api_stream_template(
  api = "strava",
  coord_system = "gps",
  token_path = "~/strava_tokens.json",
  stream_extra = c(
    velocity_smooth = "velocity_smooth",
    heartrate       = "heartrate",
    distance        = "distance"
  )
)

run_extended <- initiate(
  session  = "18529469417",
  template = strava_template
)

run_extended
names(run_extended)
quality_report(run_extended)
```

Expected columns:

``` text
unix_time, lat, lng, altitude, velocity_smooth, heartrate, distance, is_interpolated
```

You can add other Strava streams in the same way:

```{r}
strava_template_more <- api_stream_template(
  api = "strava",
  coord_system = "gps",
  token_path = "~/strava_tokens.json",
  stream_extra = c(
    velocity_smooth = "velocity_smooth",
    heartrate       = "heartrate",
    distance        = "distance",
    cadence         = "cadence",
    watts           = "watts"
  )
)

run_more <- initiate(
  session  = "18529469417",
  template = strava_template_more
)
```

Not every activity has every stream. `watts` will be missing if power
was not recorded or estimated. Start with fields you know exist.

# 3. Saving and reusing a template

Once a template works, save it to a file:

```{r}
save_api_stream_template(
  strava_template,
  "inst/templates/strava_activity_streams.xml"
)
```

Load it later by passing the XML path directly into `initiate()`:

```{r}
run_from_template_file <- initiate(
  session  = "18529469417",
  template = "inst/templates/strava_activity_streams.xml"
)
```

The saved XML describes the API source, authentication structure, stream
mappings, and coordinate system — but no actual credentials, provided
you used `$ENV_VAR` references for `auth_value`, `oauth2_client_id`, and
`oauth2_client_secret`. The template file is safe to commit. The
credential values stay in `.Renviron` or the token JSON file, which
stays in `.gitignore`.

This makes ingestion reproducible across machines, across seasons, and
across collaborators:

```{r}
initiate(session = activity_id, template = "templates/strava_running.xml")
```

# 4. Generic REST API setup

The generic REST pathway handles any API that returns JSON and can be
mapped to a `motion_trace`. Use it when you have a URL endpoint,
optional authentication, and a JSON response containing time-series data
with identifiable time and coordinate fields.

The generic pathway requires an `api_stream_template()` where
`api = "generic"`. At minimum the template needs to answer:

| Concept | Argument | Example |
|------------------------|------------------------|------------------------|
| Which API pathway? | `api` | `"generic"` |
| GPS or local coordinates? | `coord_system` | `"gps"` |
| What URL? | `url` | `"https://.../{session}/streams"` |
| How is auth applied? | `auth_type`, `auth_value` | bearer token from env var |
| Where is the data in the JSON? | `data_path` | `"data.samples"` |
| What shape is the response? | `response_format` | `"records"`, `"columnar"`, or `"streams"` |
| Which field is time? | `stream_time` | `"timestamp"` |
| Which fields are coordinates? | `stream_latlng` or `stream_lat`/`stream_lng` | `"latlng"` |
| Which extra fields to keep? | `stream_extra` | `c(hr = "heart_rate")` |

## A note on JSON structure

`initiate()` maps JSON fields directly to columns in the `motion_trace`.
For this to work, each field must be a **flat, parallel array of equal
length** — one value per time step, in the same order as the time field.
If your API wraps data inside a nested object (for example
`response → data → samples → [...]`), use `data_path = "data.samples"`
to point `initiate()` at the right level before any field mapping
happens. Deeply nested or irregular structures (fields that appear on
some records but not others, or arrays of different lengths) will need
to be flattened outside the package before ingestion.

## Response formats

The generic reader supports three common JSON shapes.

### Records

Use `response_format = "records"` when the API returns an array of
objects:

``` json
[
  {"timestamp": "2026-05-26T12:00:00Z", "lat": 40.1, "lng": -73.9, "heart_rate": 150},
  {"timestamp": "2026-05-26T12:00:01Z", "lat": 40.1, "lng": -73.9, "heart_rate": 151}
]
```

```{r}
records_template <- api_stream_template(
  api            = "generic",
  coord_system   = "gps",
  url            = "https://example.com/api/sessions/{session}/samples",
  request_method = "GET",
  auth_type      = "bearer",
  auth_value     = "$EXAMPLE_ACCESS_TOKEN",
  response_format = "records",
  time_type      = "iso8601",
  stream_time    = "timestamp",
  stream_lat     = "lat",
  stream_lng     = "lng",
  stream_extra   = c(heartrate = "heart_rate")
)
```

### Columnar

Use `response_format = "columnar"` when the API returns named arrays:

``` json
{
  "timestamp": [1770000000, 1770000001],
  "lat": [40.1, 40.1],
  "lng": [-73.9, -73.9],
  "heart_rate": [150, 151]
}
```

```{r}
columnar_template <- api_stream_template(
  api             = "generic",
  coord_system    = "gps",
  url             = "https://example.com/api/sessions/{session}/samples",
  request_method  = "GET",
  auth_type       = "bearer",
  auth_value      = "$EXAMPLE_ACCESS_TOKEN",
  response_format = "columnar",
  time_type       = "unix",
  stream_time     = "timestamp",
  stream_lat      = "lat",
  stream_lng      = "lng",
  stream_extra    = c(heartrate = "heart_rate")
)
```

### Streams

Use `response_format = "streams"` when the API returns one object per
stream with a nested data vector. This is the Strava-like shape.

``` json
{
  "time":      {"data": [0, 1, 2]},
  "latlng":    {"data": [[40.1, -73.9], [40.1, -73.9], [40.1, -73.9]]},
  "heartrate": {"data": [150, 151, 152]}
}
```

```{r}
streams_template <- api_stream_template(
  api             = "generic",
  coord_system    = "gps",
  url             = "https://example.com/api/activities/{session}/streams",
  request_method  = "GET",
  auth_type       = "bearer",
  auth_value      = "$EXAMPLE_ACCESS_TOKEN",
  response_format = "streams",
  stream_data_key = "data",
  time_type       = "unix",
  stream_time     = "time",
  stream_latlng   = "latlng",
  stream_extra    = c(heartrate = "heartrate")
)
```

# 5. Generic REST using Strava as a case study

This is not the recommended production path for Strava, but it is a
useful way to verify that the generic REST parser is working.

First, refresh your Strava token using the package internals or by
running a normal Strava initiation. During package development, if
`get_valid_tokens()` is not exported, use
`motiongrammar:::get_valid_tokens()`.

```{r}
tokens <- jsonlite::fromJSON(path.expand("~/strava_tokens.json"))
tokens <- motiongrammar:::get_valid_tokens(tokens)

Sys.setenv(STRAVA_ACCESS_TOKEN = tokens$access_token)
```

Then create a generic template pointing to Strava's streams endpoint:

```{r}
strava_generic_template <- api_stream_template(
  api            = "generic",
  coord_system   = "gps",
  url            = "https://www.strava.com/api/v3/activities/{session}/streams",
  request_method = "GET",
  request_params = list(
    keys        = "time,latlng,altitude,velocity_smooth,heartrate,distance",
    key_by_type = "true"
  ),
  auth_type       = "bearer",
  auth_value      = "$STRAVA_ACCESS_TOKEN",
  response_format = "streams",
  stream_data_key = "data",
  time_type       = "unix",
  stream_time     = "time",
  stream_latlng   = "latlng",
  stream_altitude = "altitude",
  stream_extra    = c(
    velocity_smooth = "velocity_smooth",
    heartrate       = "heartrate",
    distance        = "distance"
  )
)
```

```{r}
run_generic_strava <- initiate(
  session  = "18529469417",
  template = strava_generic_template
)

run_generic_strava
names(run_generic_strava)
quality_report(run_generic_strava)
```

Note that Strava's stream `time` is seconds elapsed from the activity
start, not wall-clock time. The dedicated `source = "strava"` path
converts this into real Unix time by fetching the activity metadata
first. The generic test above is mainly a parser check, not preferred
production Strava ingestion.

# 6. Adapting the generic REST setup to another API

Inspect the JSON response from your API and identify:

1.  The endpoint URL.
2.  The authentication style.
3.  Whether the response is `records`, `columnar`, or `streams`.
4.  Where the actual samples live inside the JSON.
5.  Which fields correspond to time, position, and any extra signals.

## Authentication types

### Simple tokens — `"bearer"` and `"api_key_*"`

Use these when the API gives you a long-lived key or token that does not
expire, or when you manage token refresh yourself outside the package.

```{r}
# Bearer token — most common; sent as Authorization: Bearer <token>
api_stream_template(
  api        = "generic",
  url        = "https://api.example.com/sessions/{session}",
  stream_time = "timestamp", stream_lat = "lat", stream_lng = "lng",
  auth_type  = "bearer",
  auth_value = "$MY_ACCESS_TOKEN"
)

# API key in a request header (e.g. X-Api-Key: <key>)
api_stream_template(
  api         = "generic",
  url         = "https://api.example.com/sessions/{session}",
  stream_time = "timestamp", stream_lat = "lat", stream_lng = "lng",
  auth_type   = "api_key_header",
  auth_header = "x-api-key",
  auth_value  = "$MY_API_KEY"
)

# API key as a URL query parameter (e.g. ?api_key=<key>)
api_stream_template(
  api         = "generic",
  url         = "https://api.example.com/sessions/{session}",
  stream_time = "timestamp", stream_lat = "lat", stream_lng = "lng",
  auth_type   = "api_key_query",
  auth_param  = "api_key",
  auth_value  = "$MY_API_KEY"
)
```

**Environment variable references**: any `auth_value` starting with `$`
is treated as an environment variable name. `auth_value = "$MY_TOKEN"`
means "read `Sys.getenv('MY_TOKEN')` at runtime." The credential never
appears in the script or the saved XML template. Set the variable in
your `.Renviron` file or before running:

``` r
Sys.setenv(MY_TOKEN = "paste_your_token_here")
```

Long-lived API keys from sports science platforms (Smartabase, Catapult
Cloud) work well with these simple patterns.

### Automatic token refresh — `"oauth2_refresh"`

Consumer platforms like Strava, Oura, Garmin, and WHOOP use **OAuth
2.0**. The key difference from a simple API key is that OAuth access
tokens expire after a short window (typically 6 hours), so anything that
runs more than once needs to refresh them automatically.

Use `auth_type = "oauth2_refresh"` when the API documentation mentions
"refresh tokens" or an OAuth 2.0 flow, when your token stops working
after a few hours, or when you see `401 Unauthorized` errors during
batch runs.

**What the fields mean, in plain English:**

| Field | What it is | Example |
|------------------------|------------------------|------------------------|
| `oauth2_token_url` | The API endpoint that issues a new access token | `"https://oura.api.com/oauth/token"` |
| `oauth2_token_path` | A local JSON file where your tokens are stored and auto-refreshed | `"~/oura_tokens.json"` |
| `oauth2_client_id` | Your application's identifier — the API provider gives you this when you register your app on their developer portal | `"$OURA_CLIENT_ID"` |
| `oauth2_client_secret` | Your application's password — kept private, never shared | `"$OURA_CLIENT_SECRET"` |

The **client ID** and **client secret** are not your personal account
login. They identify the *application* you registered with the
provider's developer portal. When you visit Strava, Oura, or Garmin's
developer pages and create an "app", they give you these two values for
that app. Store them in `.Renviron` and reference them as
`"$OURA_CLIENT_ID"` rather than pasting them into code.

The **token file** holds the current access token, its expiry time, and
the refresh token. The package reads this file, checks whether the
access token has expired, and — if it has — silently calls
`oauth2_token_url` with the refresh token to get a fresh one. The file
is created once (manually, during initial authorisation) and then
maintained automatically. A minimal token file:

``` json
{
  "access_token":  "eyJhbGci...",
  "refresh_token": "def502...",
  "expires_at":    1748000000
}
```

`expires_at` is a Unix timestamp (seconds since 1970-01-01). You do not
need to update this manually.

**Security**: add the token JSON file to `.gitignore`. It contains live
credentials.

**Example: Oura ring activity data**

```{r}
oura_template <- api_stream_template(
  api            = "generic",
  coord_system   = "local",   # Oura has no GPS — accelerometer data only
  url            = "https://api.ouraring.com/v2/usercollection/daily_activity",
  request_method = "GET",
  request_params = list(start_date = "{session}", end_date = "{session}"),

  auth_type            = "oauth2_refresh",
  oauth2_token_url     = "https://api.ouraring.com/oauth/token",
  oauth2_token_path    = "~/oura_tokens.json",
  oauth2_client_id     = "$OURA_CLIENT_ID",
  oauth2_client_secret = "$OURA_CLIENT_SECRET",

  response_format = "records",
  data_path       = "data",
  time_type       = "iso8601",
  stream_time     = "timestamp",
  stream_x        = "active_calories",
  stream_y        = "steps",
  stream_extra    = c(score = "score", met = "met_alert")
)
```

**Example: WHOOP recovery data**

```{r}
whoop_template <- api_stream_template(
  api            = "generic",
  coord_system   = "local",
  url            = "https://api.prod.whoop.com/developer/v1/cycle",
  request_method = "GET",

  auth_type            = "oauth2_refresh",
  oauth2_token_url     = "https://api.prod.whoop.com/oauth/oauth2/token",
  oauth2_token_path    = "~/whoop_tokens.json",
  oauth2_client_id     = "$WHOOP_CLIENT_ID",
  oauth2_client_secret = "$WHOOP_CLIENT_SECRET",

  response_format = "records",
  data_path       = "records",
  time_type       = "iso8601",
  stream_time     = "start",
  stream_x        = "score.strain",
  stream_y        = "score.kilojoule",
  stream_extra    = c(recovery = "score.recovery_score", hrv = "score.hrv_rmssd_milli")
)
```

Save the template and future runs are fully automated — the package
reads the template, loads the token file, refreshes if needed, and
fetches the data with no interactive login required.

```{r}
save_api_stream_template(whoop_template, "templates/whoop_cycles.xml")

# Later, in a batch run:
cycle <- initiate(session = "2026-05-28", template = "templates/whoop_cycles.xml")
```

## Time handling — `time_type`

| `time_type` | When to use | What happens |
|------------------------|------------------------|------------------------|
| `"unix"` | API returns seconds since 1970-01-01 | Values used directly as `unix_time` |
| `"iso8601"` | API returns strings like `"2026-05-28T10:00:00Z"` | Parsed and converted to Unix time |
| `"relative"` | API returns elapsed seconds from a reference point (`0, 1, 2, 3...`) | Added to `time_origin` to produce Unix time |

`"relative"` is common for wearables, IMUs, and devices like Catapult
that record session-elapsed time rather than wall-clock time. Use
`time_origin` to anchor the sequence to a real timestamp:

```{r}
catapult_template <- api_stream_template(
  api            = "generic",
  coord_system   = "local",
  url            = "https://connect.catapultsports.com/api/v6/activities/{session}/params",
  request_method = "GET",
  auth_type      = "bearer",
  auth_value     = "$CATAPULT_TOKEN",

  response_format = "records",
  time_type       = "relative",
  time_origin     = 1748430000,   # Unix timestamp for the session start
  stream_time     = "time",       # field contains 0, 1, 2, 3 ...
  stream_x        = "player_load_cumulative",
  stream_y        = "metabolic_power",
  stream_extra    = c(heart_rate = "heart_rate", speed = "speed")
)
```

If `time_origin` is omitted it defaults to `0`, meaning elapsed seconds
are used as-is for `unix_time`. This is supported within the pipeline
but will cause problems if you try to join this data to other sources
using real timestamps.

## Reserved column names

The pipeline reserves a set of column names that are overwritten by
`derivate()` and `designate()`:

```         
distance, velocity, acceleration, heading, angular_velocity,
jerk, lateral_g, designation
```

If your API returns a field that maps to one of these names in
`stream_extra` or `col_extra`, the package warns you at execution time:

```         
Warning: External field 'distance' conflicts with a reserved motiongrammar
pipeline output. derivate() / designate() will overwrite it. Consider
renaming to 'api_distance' in your template's stream_extra.
```

Use a prefixed output name to preserve the device measurement alongside
the kinematically-derived one:

```{r}
# Instead of:
stream_extra = c(distance = "distance")         # overwritten by derivate()

# Use a prefixed name:
stream_extra = c(api_distance = "distance")      # preserved through the pipeline
stream_extra = c(strava_distance = "distance")
stream_extra = c(device_velocity = "velocity_smooth")
```

The left-hand name is what the column will be called in your
`motion_trace`; the right-hand name is what the API field is called.

# 7. Google Takeout

Google Takeout is an exported archive, not a live REST API, so it is
better treated as a file-import problem.

If the export contains `.gpx` files:

```{r}
takeout_gpx <- initiate(
  source  = "gpx",
  session = "path/to/takeout/activity.gpx"
)
```

If the export contains CSV files:

```{r}
takeout_csv <- initiate(
  source  = "guess_csv",
  session = "path/to/takeout/location_history.csv"
)
```

If the export contains JSON files, that does not currently fit the
generic REST pathway (which expects a live URL). Conceptually the
mapping is the same — identify the time field, coordinates, and extra
streams. A future extension could add `source = "json_template"` or
allow the generic API template to read from a local JSON file, which
would also cover Apple Health exports and other archived formats.

# 8. Production summary

-   Use `source = "strava"` for straightforward Strava ingestion.
-   Use `api_stream_template(api = "strava", ...)` when you need extra
    streams alongside the default spatial data.
-   Save stable templates with `save_api_stream_template()` and load
    them by passing the file path to `initiate()`. Templates can be
    committed to version control safely as long as credentials use
    `$ENV_VAR` references.
-   Use `api = "generic"` for any other JSON REST API.
-   Store all credentials — API keys, client secrets, token JSON files —
    outside your scripts. `.Renviron` for keys, `.gitignore` for token
    files. Never commit either to a repository.
