---
title: "Interpolate: Repairing Gaps in Tracking Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interpolate: Repairing Gaps in Tracking Data}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4,
  warning  = FALSE,
  message  = FALSE
)
```

```{r pipeline-setup, include=FALSE}
library(motiongrammar)
library(ggplot2)
library(dplyr)

# Build trace up to the point before interpolation
trace_pre <- initiate(
  session = system.file("extdata", "example_gps_file_short.csv", package = "motiongrammar"),
  source  = 'guess_csv'
) |>
  coordinate(from = "latlong", to = "xyz", crs = 3857, norm = TRUE,
             drop_outliers = TRUE)
```

`interpolate()` repairs gaps in motion data by constructing a dense and
evenly spaced time grid at the target frame rate. In other words, we use
this step to fill missing positions.

------------------------------------------------------------------------

## 1. Why we interpolate

When we are filtering or calculating derivatives like velocity, we
assume our data is evenly spaced (ie., there is one row every 10th of a
second for every coordinate). Unfortunately, devices drop out so this is
not always true Interpolation allows us to make an educated guess on
what the coordinate *should be* when drop outs occur. This is another
thing that is usually done by providers, but is useful to check when
publishing a paper.

------------------------------------------------------------------------

## 2. Basic usage

Here, you just need to set the interpolation method (explained more
depth below), the sample rate and the maximum gap in frames. We set a
max gap in frames, because sometimes dropouts can be *so extreme* that
our ability to make an educated guess becomes questionable. In the
example below, we fill any drop outs less than 20 samples (2 seconds,
because 10 Hz x 2 seconds = 20). But if there was a drop out of, for
example, ten thousand seconds, then we truly have no idea where the
player was in that time. All interpolation methods can only connect
"when we saw the player last" to "when we see them again".

```{r interpolate}
trace <- trace_pre |>
  interpolate(
    method         = "linear",
    hz             = 10,
    max_gap_frames = 20    # fill gaps up to 2 seconds; leave larger ones as NA
  )

```

------------------------------------------------------------------------

## 3. Visualising the `is_interpolated` flag

Every filled row receives `is_interpolated = TRUE`. You can see exactly
which positions are real and which are estimated. In our example, there
is nothing to estimate, as it has been done by the provider.

```{r plot-interpolated}
trace |>
  filter(!is.na(x)) |>
  ggplot(aes(x = x, y = y, colour = is_interpolated)) +
  geom_point(size = 0.4, alpha = 0.6) +
  coord_fixed() +
  scale_colour_manual(
    values = c("FALSE" = "#2B7C85", "TRUE" = "#FF6F59"),
    labels = c("FALSE" = "Recorded", "TRUE" = "Interpolated")
  ) +
  labs(
    title  = "Recorded vs interpolated positions",
    x      = "x (m)", y = "y (m)", colour = NULL
  ) +
  theme_minimal() +
  theme(panel.grid = element_blank(), legend.position = "top")
```

In a well-functioning GPS session most points should be blue. Clusters
of red indicate where the signal dropped and linear fill was used.

A spatial plot shows *where* gaps occurred. A time-series plot shows
*when* they occurred and how the filled values connect to the
surrounding real data:

```{r plot-interpolated-time}
trace |>
  filter(!is.na(x)) |>
  ggplot(aes(x = unix_time, y = x)) +
  geom_line(colour = "#2B7C85", linewidth = 0.4) +
  geom_point(
    data  = \(d) filter(d, is_interpolated),
    colour = "#FF6F59", size = 1.2
  ) +
  labs(
    title    = "Visualising repaired data gaps",
    subtitle = "Orange points = interpolated; line = full reconstructed trajectory",
    x = "Time (Unix seconds)", y = "x (m)"
  ) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank())
```

Each orange point is a position the package estimated rather than
recorded. A short isolated cluster means a brief signal hiccup — fine
to keep. A long run of orange covering a sharp directional change is
worth scrutinising before running kinematics.

------------------------------------------------------------------------

## 4. The `max_gap_frames` threshold

Not all gaps should be filled. A 0.3-second GPS dropout during a sprint
is sensor noise — interpolating it is reasonable. A 600-second gap while
the player was indoors is more egregious. We cannot make an educated
guess on where a player would be during this time, so filling would be
fabricating data.

`max_gap_frames` (default: 5) controls this boundary. Gaps of this many
consecutive missing frames or fewer are filled; longer gaps are left as
`NA`.

```{r max-gap, eval=FALSE}
# At 10 Hz: fill gaps up to 2 seconds; leave longer gaps as NA
trace <- trace_pre |>
  interpolate(hz = 10, max_gap_frames = 20)

# More conservative: only fill single-frame dropouts
trace <- trace_pre |>
  interpolate(hz = 10, max_gap_frames = 3)
```

**How to set it:** Multiply your threshold in seconds by `hz`. For
field-sport GPS, 1–3 seconds (10–30 frames at 10 Hz) is typical.
Anything longer is likely a genuine off-field period that `designate()`
handles better than interpolation.

------------------------------------------------------------------------

## 5. Interpolation methods

### `"linear"` (default)

Draws a straight line between the last known position and the first
known position after the gap. Safe for short dropouts.

### `"spline"`

Fits a natural cubic spline through the surrounding valid positions.
Better at following curved paths, but can overshoot in noisy data.
Theoretically, this is a better match for heavy CoD sports, but I would
always check by visualising.

### `"constant"` — last observation carried forward

Repeats the last known position for every missing frame. Treats the
player as stationary during the gap. Useful for RFID / UWB systems with
zone-boundary dropouts.

```{r methods-compare}
trace_spline   <- trace_pre |> interpolate(method = "spline",   hz = 10, max_gap_frames = 20)
trace_constant <- trace_pre |> interpolate(method = "constant", hz = 10, max_gap_frames = 20)

# Compare: how many NAs remain?
lapply(list(linear = trace, spline = trace_spline, constant = trace_constant), function(tr) {
  c(na_x = sum(is.na(tr$x)), interpolated = sum(tr$is_interpolated, na.rm = TRUE))
})
```

------------------------------------------------------------------------

## 6. Reading the quality log

```{r quality}
quality_report(trace,
               step = 'interpolate')
```

The interpolate section shows the gap structure before filling (count,
min/max/mean gap length, how many exceeded `max_gap_frames`), how many
NAs were filled in x and y, and how many remain. A high interpolation
ratio (\> 20–30%) is flagged as an issue.

------------------------------------------------------------------------

## Quick reference

| Argument | Default | Purpose |
|----|----|----|
| `method` | `"linear"` | `"linear"`, `"spline"`, or `"constant"` |
| `hz` | from metadata | Target frame rate |
| `max_gap_frames` | `5` | Gaps larger than this (frames) are left as `NA` |
| `fill_passthrough` | `"none"` | `"constant"` forward-fills non-numeric columns (e.g. zone labels) through inserted rows |

**Remaining NAs after interpolation are intentional.** If you still see
`NA` in `x` or `y` after calling `interpolate()`, it means those gaps
were wider than `max_gap_frames` and were deliberately left unfilled.
This is the safety threshold working as designed — the function is
protecting you from fabricating positions across gaps so large that any
estimated path would be meaningless. To fill more gaps, increase
`max_gap_frames`; to understand which gaps were skipped, check
`quality_report(trace, step = "interpolate")` and look at
`gaps$gaps_exceeding_maxgap`.
