---
title: "Interpolate: Repairing Gaps in Tracking Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interpolate: Repairing Gaps in Tracking Data}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4,
  warning  = FALSE,
  message  = FALSE
)
```

```{r pipeline-setup, include=FALSE}
devtools::load_all()
library(ggplot2)
library(dplyr)

# Build trace up to the point before interpolation
trace_pre <- initiate(
  session = '~/motiongrammar/data/example_gps_file_short.csv',
  source  = 'catapult_replay'
) |>
  coordinate(from = "latlong", to = "xyz", crs = 3857, norm = TRUE,
             drop_outliers = TRUE)
```

`interpolate()` repairs the gap in our GPS data by constructing a dense,
evenly-spaced time grid at the target frame rate and filling missing
positions.

------------------------------------------------------------------------

## 1. Why you need interpolation

Most downstream calculations — filtering, differentiation, distance
accumulation — assume frames are evenly spaced in time. A raw GPS file
with irregular timing violates this assumption and produces incorrect
derivatives. Even a 1-frame gap at 10 Hz means the finite difference at
that boundary spans 0.2 s instead of 0.1 s, doubling the computed
velocity for that frame.

Our example file illustrates this clearly. The player was indoors for
the first ~10 minutes with no GPS signal — those rows arrived as lat/lng
= 0 and were set to NA by `coordinate()`:

```{r show-gap}
library(motiongrammar)
library(ggplot2)
library(dplyr)

trace_pre <- initiate(
  session = '~/motiongrammar/data/example_gps_file_short.csv',
  source  = 'catapult_replay'
) |>
  coordinate(from = "latlong", to = "xyz", crs = 3857, norm = TRUE,
             drop_outliers = TRUE)

# How many rows have no GPS?
cat("Total rows:", nrow(trace_pre), "\n")
cat("Rows with NA x (no GPS fix):", sum(is.na(trace_pre$x)), "\n")
cat("Rows with valid GPS:", sum(!is.na(trace_pre$x)), "\n")
```

```{r plot-pre-interp}
trace_pre |>
  mutate(has_gps = !is.na(x)) |>
  ggplot(aes(x = unix_time, y = ifelse(has_gps, 1, 0), fill = has_gps)) +
  geom_col(width = 1) +
  scale_fill_manual(
    values = c("FALSE" = "firebrick2", "TRUE" = "steelblue"),
    labels = c("FALSE" = "No GPS fix", "TRUE" = "Valid GPS")
  ) +
  labs(
    title = "GPS signal availability over time",
    x     = "Unix time (s)",
    y     = NULL,
    fill  = NULL
  ) +
  theme_minimal() +
  theme(
    panel.grid    = element_blank(),
    axis.text.y   = element_blank(),
    axis.ticks.y  = element_blank(),
    legend.position = "top"
  )
```

The red block at the start is the ~600-second indoor period.
`interpolate()` leaves this large gap as `NA` (it exceeds
`max_gap_frames`) and fills only the smaller dropouts within the GPS
period.

------------------------------------------------------------------------

## 2. Basic usage

```{r interpolate}
trace <- trace_pre |>
  interpolate(
    method         = "linear",
    hz             = 10,
    max_gap_frames = 20    # fill gaps up to 2 seconds; leave larger ones as NA
  )

cat("Rows before:", nrow(trace_pre), "\n")
cat("Rows after:", nrow(trace), "\n")
cat("Rows interpolated:", sum(trace$is_interpolated, na.rm = TRUE), "\n")
cat("Interpolation rate:", round(mean(trace$is_interpolated, na.rm = TRUE) * 100, 1), "%\n")
```

------------------------------------------------------------------------

## 3. Visualising the `is_interpolated` flag

Every filled row receives `is_interpolated = TRUE`. You can see exactly
which positions are real and which are estimated:

```{r plot-interpolated}
trace |>
  filter(!is.na(x)) |>
  ggplot(aes(x = x, y = y, colour = is_interpolated)) +
  geom_point(size = 0.4, alpha = 0.6) +
  coord_fixed() +
  scale_colour_manual(
    values = c("FALSE" = "steelblue", "TRUE" = "firebrick2"),
    labels = c("FALSE" = "Recorded", "TRUE" = "Interpolated")
  ) +
  labs(
    title  = "Recorded vs interpolated positions",
    x      = "x (m)", y = "y (m)", colour = NULL
  ) +
  theme_minimal() +
  theme(panel.grid = element_blank(), legend.position = "top")
```

In a well-functioning GPS session most points should be blue. Clusters of
red indicate where the signal dropped and linear fill was used.

------------------------------------------------------------------------

## 4. The `max_gap_frames` threshold

Not all gaps should be filled. A 0.3-second GPS dropout during a sprint
is sensor noise — interpolating it is reasonable. A 600-second gap while
the player was indoors is not a dropout — filling it would fabricate
positions that never existed.

`max_gap_frames` (default: 5) controls this boundary. Gaps of this many
consecutive missing frames or fewer are filled; longer gaps are left as
`NA`.

```{r max-gap, eval=FALSE}
# At 10 Hz: fill gaps up to 2 seconds; leave longer gaps as NA
trace <- trace_pre |>
  interpolate(hz = 10, max_gap_frames = 20)

# More conservative: only fill single-frame dropouts
trace <- trace_pre |>
  interpolate(hz = 10, max_gap_frames = 3)
```

**How to set it:** Multiply your threshold in seconds by `hz`.
For field-sport GPS, 1–3 seconds (10–30 frames at 10 Hz) is typical.
Anything longer is likely a genuine off-field period that `designate()`
handles better than interpolation.

------------------------------------------------------------------------

## 5. Interpolation methods

### `"linear"` (default)

Draws a straight line between the last known position and the first known
position after the gap. Safe for short dropouts.

### `"spline"`

Fits a natural cubic spline through the surrounding valid positions. Better
at following curved paths, but can overshoot in noisy data — always
check the trajectory plot after using spline.

### `"constant"` — last observation carried forward

Repeats the last known position for every missing frame. Treats the
player as stationary during the gap. Useful for RFID / UWB systems with
zone-boundary dropouts.

```{r methods-compare}
trace_spline   <- trace_pre |> interpolate(method = "spline",   hz = 10, max_gap_frames = 20)
trace_constant <- trace_pre |> interpolate(method = "constant", hz = 10, max_gap_frames = 20)

# Compare: how many NAs remain?
lapply(list(linear = trace, spline = trace_spline, constant = trace_constant), function(tr) {
  c(na_x = sum(is.na(tr$x)), interpolated = sum(tr$is_interpolated, na.rm = TRUE))
})
```

------------------------------------------------------------------------

## 6. Reading the quality log

```{r quality}
quality_report(trace)
```

The interpolate section shows the gap structure before filling (count,
min/max/mean gap length, how many exceeded `max_gap_frames`), how many
NAs were filled in x and y, and how many remain. A high interpolation
ratio (> 20–30%) is flagged as an issue.

------------------------------------------------------------------------

## Quick reference

| Argument         | Default    | Purpose                                          |
|------------------|------------|--------------------------------------------------|
| `method`         | `"linear"` | `"linear"`, `"spline"`, or `"constant"`          |
| `hz`             | from metadata | Target frame rate                             |
| `max_gap_frames` | `5`        | Gaps larger than this (frames) are left as NA    |
