---
title: "Interpolate: Repairing Gaps in Tracking Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interpolate: Repairing Gaps in Tracking Data}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4,
  warning  = FALSE,
  message  = FALSE
)
```

```{r pipeline-setup, include=FALSE}
devtools::load_all()
require(motiongrammar)
require(tidyverse)

# Build trace up to the point before interpolation
trace_pre <- initiate(
  session = '~/motiongrammar/data/example_gps_file_short.csv',
  source  = 'catapult_replay'
) |>
  coordinate(from = "latlong", to = "xyz", crs = 3857, norm = TRUE,
             drop_outliers = TRUE)
```

`interpolate()` repairs the gap in our GPS data by constructing a dense,
evenly-spaced time grid at the target frame rate and filling missing
positions.

------------------------------------------------------------------------

## 1. Why we interpolate

When we are filtering or calculating derivatives like velocity, we
assume our data is dense or evenly spaced. In other words, we expect a
position coordinate for every time stamp. Unfortunately, devices can
drop out so this is not always the case. Interpolation allows us to make
an educated guess on what the coordinate *should be* when drop outs
occur. This is another thing that is usually done by providers, but is
useful to check when publishing a paper.

------------------------------------------------------------------------

## 2. Basic usage

```{r interpolate}
trace <- trace_pre |>
  interpolate(
    method         = "linear",
    hz             = 10,
    max_gap_frames = 20    # fill gaps up to 2 seconds; leave larger ones as NA
  )

```

------------------------------------------------------------------------

## 3. Visualising the `is_interpolated` flag

Every filled row receives `is_interpolated = TRUE`. You can see exactly
which positions are real and which are estimated. In our example, there
is nothing to estimate, as it has been done by the provider.

```{r plot-interpolated}
trace |>
  filter(!is.na(x)) |>
  ggplot(aes(x = x, y = y, colour = is_interpolated)) +
  geom_point(size = 0.4, alpha = 0.6) +
  coord_fixed() +
  scale_colour_manual(
    values = c("FALSE" = "steelblue", "TRUE" = "firebrick2"),
    labels = c("FALSE" = "Recorded", "TRUE" = "Interpolated")
  ) +
  labs(
    title  = "Recorded vs interpolated positions",
    x      = "x (m)", y = "y (m)", colour = NULL
  ) +
  theme_minimal() +
  theme(panel.grid = element_blank(), legend.position = "top")
```

In a well-functioning GPS session most points should be blue. Clusters
of red indicate where the signal dropped and linear fill was used.

------------------------------------------------------------------------

## 4. The `max_gap_frames` threshold

Not all gaps should be filled. A 0.3-second GPS dropout during a sprint
is sensor noise — interpolating it is reasonable. A 600-second gap while
the player was indoors is more egregious. We cannot make an educated
guess on where a player would be during this time, so filling would be
fabricating data.

`max_gap_frames` (default: 5) controls this boundary. Gaps of this many
consecutive missing frames or fewer are filled; longer gaps are left as
`NA`.

```{r max-gap, eval=FALSE}
# At 10 Hz: fill gaps up to 2 seconds; leave longer gaps as NA
trace <- trace_pre |>
  interpolate(hz = 10, max_gap_frames = 20)

# More conservative: only fill single-frame dropouts
trace <- trace_pre |>
  interpolate(hz = 10, max_gap_frames = 3)
```

**How to set it:** Multiply your threshold in seconds by `hz`. For
field-sport GPS, 1–3 seconds (10–30 frames at 10 Hz) is typical.
Anything longer is likely a genuine off-field period that `designate()`
handles better than interpolation.

------------------------------------------------------------------------

## 5. Interpolation methods

### `"linear"` (default)

Draws a straight line between the last known position and the first
known position after the gap. Safe for short dropouts.

### `"spline"`

Fits a natural cubic spline through the surrounding valid positions.
Better at following curved paths, but can overshoot in noisy data.
Theoretically, this is a better match for heavy CoD sports, but I would
always check by visualising.

### `"constant"` — last observation carried forward

Repeats the last known position for every missing frame. Treats the
player as stationary during the gap. Useful for RFID / UWB systems with
zone-boundary dropouts.

```{r methods-compare}
trace_spline   <- trace_pre |> interpolate(method = "spline",   hz = 10, max_gap_frames = 20)
trace_constant <- trace_pre |> interpolate(method = "constant", hz = 10, max_gap_frames = 20)

# Compare: how many NAs remain?
lapply(list(linear = trace, spline = trace_spline, constant = trace_constant), function(tr) {
  c(na_x = sum(is.na(tr$x)), interpolated = sum(tr$is_interpolated, na.rm = TRUE))
})
```

------------------------------------------------------------------------

## 6. Reading the quality log

```{r quality}
quality_report(trace)
```

The interpolate section shows the gap structure before filling (count,
min/max/mean gap length, how many exceeded `max_gap_frames`), how many
NAs were filled in x and y, and how many remain. A high interpolation
ratio (\> 20–30%) is flagged as an issue.

------------------------------------------------------------------------

## Quick reference

| Argument | Default | Purpose |
|----|----|----|
| `method` | `"linear"` | `"linear"`, `"spline"`, or `"constant"` |
| `hz` | from metadata | Target frame rate |
| `max_gap_frames` | `5` | Gaps larger than this (frames) are left as NA |
