---
title: "Quantitate: Summarising a Motion Trace"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quantitate: Summarising a Motion Trace}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4,
  warning  = FALSE,
  message  = FALSE
)
```

```{r pipeline-setup, include=FALSE}
devtools::load_all()
library(ggplot2)
library(dplyr)

# Full pipeline up to allocate — we demonstrate quantitate() here
trace <- initiate(
  session = '~/motiongrammar/data/example_gps_file_short.csv',
  source  = 'catapult_replay'
) |>
  coordinate(from = "latlong", to = "xyz", crs = 3857, norm = TRUE,
             drop_outliers = TRUE) |>
  interpolate(hz = 10, max_gap_frames = 20) |>
  filtrate(method = "butterworth", target = "coordinates", cutoff = 1.5) |>
  derivate(derivatives = c("velocity", "acceleration", "angular_velocity")) |>
  designate(method = "corbett", v_threshold = 0.3) |>
  allocate(
    allocation = list(
      velocity     = c(0, 2, 4, 6, 8),
      acceleration = c(0, 1, 2, 3, 4)
    ),
    allocation_name = "gps_standard"
  )
```

`quantitate()` collapses a frame-by-frame `motion_trace` into a tidy
summary table grouped by designation segment, intensity band, or both.
You define what to compute — distance, time, peak velocity — and it
groups and summarises accordingly.

------------------------------------------------------------------------

## 1. The three scopes

### `scope = 'designation'` — per-segment breakdown

```{r load-full-pipeline, include=FALSE}
library(motiongrammar)
library(ggplot2)
library(dplyr)

trace <- initiate(
  session = '~/motiongrammar/data/example_gps_file_short.csv',
  source  = 'catapult_replay'
) |>
  coordinate(from = "latlong", to = "xyz", crs = 3857, norm = TRUE,
             drop_outliers = TRUE) |>
  interpolate(hz = 10, max_gap_frames = 20) |>
  filtrate(method = "butterworth", target = "coordinates", cutoff = 1.5) |>
  derivate(derivatives = c("velocity", "acceleration", "angular_velocity")) |>
  designate(method = "corbett", v_threshold = 0.3) |>
  allocate(
    allocation = list(
      velocity     = c(0, 2, 4, 6, 8),
      acceleration = c(0, 1, 2, 3, 4)
    ),
    allocation_name = "gps_standard"
  )
```

```{r by-segment}
library(motiongrammar)
library(ggplot2)
library(dplyr)

summary_by_segment <- trace |>
  quantitate(
    allocation  = "gps_standard",
    derivative  = "velocity",
    scope       = "designation",
    active_only = FALSE,
    n_frames    = dplyr::n(),
    time_s      = dplyr::n() / 10,
    dist_m      = sum(distance, na.rm = TRUE),
    mean_vel    = round(mean(velocity, na.rm = TRUE), 2),
    peak_vel    = round(max(velocity, na.rm = TRUE), 2)
  )

summary_by_segment
```

### `scope = 'session'` — whole-session band totals

```{r by-session}
session_totals <- trace |>
  quantitate(
    allocation = "gps_standard",
    derivative = "velocity",
    scope      = "session",
    time_s     = dplyr::n() / 10,
    dist_m     = sum(distance, na.rm = TRUE),
    peak_vel   = round(max(velocity, na.rm = TRUE), 2)
  )

session_totals
```

### `scope = 'active'` — active periods only

```{r by-active}
active_totals <- trace |>
  quantitate(
    allocation = "gps_standard",
    derivative = "velocity",
    scope      = "active",
    time_s     = dplyr::n() / 10,
    dist_m     = sum(distance, na.rm = TRUE)
  )

active_totals
```

------------------------------------------------------------------------

## 2. Visualising the summary

### Distance by velocity band

```{r plot-dist-band}
session_totals |>
  filter(!is.na(gps_standard_velocity_band)) |>
  ggplot(aes(x = gps_standard_velocity_band, y = dist_m,
             fill = gps_standard_velocity_band)) +
  geom_col(show.legend = FALSE) +
  scale_fill_viridis_d(option = "plasma", begin = 0.1, end = 0.9) +
  labs(
    title = "Distance covered in each velocity band (whole session)",
    x     = "Velocity band",
    y     = "Distance (m)"
  ) +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

### Time by velocity band

```{r plot-time-band}
session_totals |>
  filter(!is.na(gps_standard_velocity_band)) |>
  ggplot(aes(x = gps_standard_velocity_band, y = time_s,
             fill = gps_standard_velocity_band)) +
  geom_col(show.legend = FALSE) +
  scale_fill_viridis_d(option = "plasma", begin = 0.1, end = 0.9) +
  labs(
    title = "Time in each velocity band (whole session)",
    x     = "Velocity band",
    y     = "Duration (s)"
  ) +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

### Designation × band heatmap

```{r plot-heatmap, fig.height=3}
summary_by_segment |>
  filter(!is.na(gps_standard_velocity_band)) |>
  ggplot(aes(x = gps_standard_velocity_band, y = designation,
             fill = dist_m)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  scale_fill_viridis_c(option = "plasma", begin = 0.1, end = 0.9) +
  labs(
    title  = "Distance (m) by segment and velocity band",
    x      = "Velocity band",
    y      = "Designation",
    fill   = "Dist (m)"
  ) +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

------------------------------------------------------------------------

## 3. Summarising without intensity bands

```{r no-bands}
# One row per designation segment
segment_totals <- trace |>
  quantitate(
    allocation = NULL,
    scope      = "designation",
    time_s     = dplyr::n() / 10,
    dist_m     = sum(distance, na.rm = TRUE),
    peak_vel   = round(max(velocity, na.rm = TRUE), 2)
  )

segment_totals
```

------------------------------------------------------------------------

## 4. `band_duration()` — convenience wrapper

```{r band-duration}
duration_tbl <- band_duration(
  trace,
  allocation  = "gps_standard",
  derivative  = "velocity",
  scope       = "designation",
  active_only = TRUE
)

duration_tbl
```

```{r plot-duration}
duration_tbl |>
  filter(!is.na(gps_standard_velocity_band)) |>
  ggplot(aes(x = gps_standard_velocity_band, y = duration_sec,
             fill = gps_standard_velocity_band)) +
  geom_col(show.legend = FALSE) +
  scale_fill_viridis_d(option = "plasma", begin = 0.1, end = 0.9) +
  labs(
    title = "Duration in each velocity band (active segments only)",
    x     = "Velocity band", y = "Duration (s)"
  ) +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

------------------------------------------------------------------------

## 5. Saving results

```{r save, eval=FALSE}
# Preserves the full quality log
saveRDS(summary_by_segment, "~/outputs/session_summary.rds")

# CSV — attributes are dropped (expected)
readr::write_csv(summary_by_segment, "~/outputs/session_summary.csv")

# Save quality log separately if needed alongside CSV
saveRDS(attr(summary_by_segment, "quality"), "~/outputs/session_quality.rds")
```

------------------------------------------------------------------------

## 6. Reading the quality log

```{r quality}
qual <- attr(summary_by_segment, "quality")
names(qual)   # every step that ran
qual$quantitate[[1]]$scope
qual$quantitate[[1]]$group_vars
qual$quantitate[[1]]$n_rows_in
qual$quantitate[[1]]$n_rows_out
```

The quality log attached to the summary tibble contains the complete
pipeline record — from `initiate()` through to `quantitate()`.

------------------------------------------------------------------------

## Quick reference

| Argument      | Default          | Purpose                                              |
|---------------|------------------|------------------------------------------------------|
| `allocation`  | `"allocation_1"` | Allocation name, or `NULL` for no band grouping      |
| `derivative`  | `"velocity"`     | Which band columns to group by (one or more)         |
| `scope`       | `"designation"`  | `"designation"`, `"session"`, or `"active"`          |
| `active_only` | `FALSE`          | Restrict to `active_*` segments (designation scope)  |
| `...`         | —                | Any `dplyr::summarise()` expressions                 |
