---
title: "motiongrammar Quickstart"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{motiongrammar Quickstart}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  warning = FALSE,
  message = FALSE
)
```

This vignette walks through the full **motiongrammar** pipeline — from a
raw CSV of latitude and longitude, to a cleaned, filtered, segmented,
and summarised motion trace.

The pipeline reads left-to-right:

```         
initiate → coordinate → interpolate → filtrate → derivate → filtrate (derivatives)
         → designate → elaborate → allocate → quantitate
```

------------------------------------------------------------------------

## 0. Setup

First we need to install motiongrammar. This package is not on CRAN, so
we run the following (only need to do this once):

```{r}
devtools::load_all()
```

First, let's load in the packages we will use for this vignette.

```{r load}
require(motiongrammar)
require(tidyverse)
```

------------------------------------------------------------------------

## 1. Initiate — load the CSV

*motiongrammar* is compatible with:

-   csv files

-   gpx files

-   API streams

For our quickstart, however, let's use a synthetic GPS file to
demonstrate the "guess" functionality of *motiongrammar*. The guess
algorithm:

1.  determines where timestamp, latitude and longitude lie
2.  Cleans out any additional rows or columns

In a prod workflow, we would use the template functionality (covered
elsewhere) to ensure everything is read in a standardised way. The
template method also allows the user to save custom extensions, for
example PlayerLoad or MetabolicPower, which by default are ignored by
the *guess* method.

```{r initiate}
trace <- initiate(session = '~/motiongrammar/data/example_gps_file_short.csv',
                  method = 'guess')

```

Now we have initiated a motiongrammar object. This is comprised of:

1.  The data itself – unix_time, lat, lng, altitude, is_interpolayed.
2.  metadata - rather than storing metadata in every column, we save it
    as a memory efficient object. An example of this is shown below. We
    can access this using "metadata_report"

```{r metadata}
metadata_report(trace)
```

3.  quality - in academic settings, we want to preserve every
    transformation done on the dataset. *quality* does this
    automatically, and populates with how the data has changed at every
    step in *motiongrammar*. It also flags any major red flags in the
    data set. We can access this using "quality_report"

```{r quality }
quality_report(trace)
```

------------------------------------------------------------------------

## 2. Coordinate — project lat/lng to metres

GPS data is spherical (degrees). `coordinate()` projects it to a flat
Cartesian plane so that distances, velocities, and headings are in
metres and m/s.

`norm = TRUE` shifts the origin so the player's first recorded position
is (0, 0). This makes trajectories easier to plot and interpret.
`drop_outliers = TRUE`drops data points well outside of range of the
rest of the data. So for example, if a player goes into the locker room
at half time.

This step also supports conversions between metric and imperial systems
(but by default, we use the SI metric).

```{r coordinate}
trace <- trace |>
  coordinate(from = "latlong",
    to = "xyz",
    crs = 3857,
    norm = TRUE,
    drop_outliers = TRUE)
```

Let's do a quick plot of our data as a sanity check:

```{r}
trace |> 
ggplot(aes(x = x, y = y)) +
  geom_path(alpha = 0.4, 
            linewidth = 0.4) +
  coord_fixed() +
  labs(title = "Raw projected trajectory (before interpolation)", 
       x = "x (m)",
       y = "y (m)") +
  theme_minimal() +
  theme(panel.grid = element_blank())

```

------------------------------------------------------------------------

## 3. Interpolate — fill dropouts at the target frame rate

GPS data drops out - interpolate handles this, if the provider has not
already! `interpolate()` builds a dense time grid at `hz` frames per
second and fills any gaps shorter than `max_gap_frames` using the chosen
method. The logic here is that, a drop out above max_gap_frames is not
just poor signal, but the battery dying (or the player being in the
locker room for example)

```{r interpolate}
trace <- trace |>
  interpolate(
    method = "linear",
    hz = 10,
    max_gap_frames = 20  # gaps > 2 s are left as NA
  )
```

------------------------------------------------------------------------

## 4.0 Finding the cutoff frequency — coordinates

Most providers handle filtering for us. But in an academic setting, this
is something we need to check and sometimes do ourselves.
*motiongrammar* includes functionality to easily identify cutoff
frequency for a low pass butterworth filter using two methods:

**Residual analysis** (Winter, 2009): sweeps a range of cutoffs, applies
the filter at each, and plots the RMS of the residuals (raw − filtered).
The elbow of the curve marks where adding more smoothing removes real
signal rather than noise.

**Power Spectral Density (PSD)**: shows how power is distributed across
frequencies. Signal energy typically concentrates below a threshold;
noise is relatively flat above it.

```{r cutoff-coordinates, fig.height=5}
filtrate_cutoff(
  trace,
  method = "butterworth",
  target = "coordinates",
  axis = "x", # analyse the x axis; repeat for y if desired
  plot_type = "both", # show residual curve and PSD side-by-side
  cutoff_range = c(0.1, 4),
  n_steps = 60
)
```

The dashed red line is the suggested cut off (maximum curvature of the
residual curve). Use it as a starting point and adjust based on the PSD.

A cutoff of **\~1–2 Hz** is typical for GPS-derived position data at 10
Hz. Higher cutoffs retain more of the true path shape; lower cutoffs are
appropriate when GPS noise is severe.

------------------------------------------------------------------------

## 4.5 Filtrate — smooth the coordinates

Now we actually apply a 2nd-order zero-phase Butterworth low-pass filter
at the chosen cutoff. The filtered coordinates are stored in new `f_x`,
`f_y` columns; `derivate()` will use these by default.

```{r filtrate-coordinates}
trace <- trace |>
  filtrate(method = "butterworth",
    target = "coordinates",
    n = 2, # filter order
    cutoff = 1.5, # Hz — chosen from the analysis above
    type = "low")

```

Below, we compare filtered versus recorded data. This data has clearly
already been filtered by the provider, so we see minimal differences
between filtered versus actual.

```{r}
# Compare raw vs filtered on a short window
trace |> 
  drop_na(f_x) |> 
ggplot(aes(x = unix_time)) +
  geom_line(aes(y = x, 
                colour = "Raw")) +
  geom_line(aes(y = f_x,
                colour = "Filtered")) +
  scale_colour_manual(values = c(Raw = "firebrick2", 
                                 Filtered = "dodgerblue")) +
  labs(title = "x coordinate - filtered versus actual",
       x = "Unix time (s)", 
       y = "x (m)", 
       colour = NULL) +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

------------------------------------------------------------------------

## 5. Derivate — velocity, acceleration, angular velocity

`derivate()` computes derivatives using central differences over a
rolling window. By default it uses the filtered `f_x` / `f_y` columns.

Velocity and acceleration are fairly standard here, but angular_velocity
is something I wanted to add leading on from Alice Sweeting's PhD. It
provides a continuous time-series of whether a player is running in a
straight or curved line.

```{r derivate}
trace <- trace |>
  derivate(use_filtered = TRUE,
    derivatives = c("velocity", 
                    "acceleration", 
                    "angular_velocity")
  )

summary(trace[, c("velocity", "acceleration", "angular_velocity")])
```

We can look at angular velocity as a time-series

```{r ts_angularvelocity}
trace |> 
  drop_na(angular_velocity) |> 
ggplot(aes(x = unix_time, 
           y = angular_velocity)) +
  geom_line(linewidth = 0.3, 
            colour = "black") +
  labs(title = "Angular Velocity", 
       x = "Time (s)", 
       y = "Angular Velocity (deg/s)") +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

Or as co-ordinates. Here, notice how when a player changes direction, it
is picked up by angular velocity:

```{r field_angularvelocity}
trace |> 
  drop_na(angular_velocity) |> 
ggplot(aes(x = f_x,
           y = f_y,
           color = angular_velocity,
           fill = angular_velocity)) +
  geom_path(linewidth = 1.5) +
  labs(title = "Angular Velocity", 
       x = "filtered x", 
       y = "filtered y") +
  theme_minimal() +
  theme(panel.grid = element_blank())+
    scale_color_viridis_c(option = "viridis",
    begin = 0.1,
    end   = 0.9)
```

------------------------------------------------------------------------

## 6. If desired, you can filter derivatives as opposed to coordinates.

Whether to filter on a coordinates or derivatives is a debate in
biomechanics, so I decided to offer both. So let's take a brief
interlude to look at how we might determine cut off frequency on
velocity instead of co-ordinates.

```{r cutoff-velocity, fig.height=5}
filtrate_cutoff(trace,
  method = "butterworth",
  target = "velocity",
  plot_type = "both",
  cutoff_range = c(0.1, 4),
  n_steps = 60)
```

## 7. Designate — segment activity vs rest

`designate()` allows us to split an activity into drills, quarters etc.
In *motiongrammar*, I have included a native PELT changepoint detection
on the velocity signal to identify contiguous activity bouts and relief
periods.

```{r designate}
trace <- trace |>
  designate(v_threshold = 0.3,
    method = "corbett")
```

Visualise the segmentation.

```{r designate-plot}
trace |> 
ggplot(aes(x = unix_time, 
           y = velocity, 
           colour = designation)) +
  geom_line(linewidth = 0.4) +
  labs(title = "Velocity coloured by designation",
       x = "Time (s)", 
       y = "Velocity (m/s)", 
       colour = "Segment") +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        legend.position = "bottom")
```

In this example, we only end up with one segment. This is due to the
synthetic data used – there literally is only one period here. Future
versions of *motiongrammar* will incorporate spatial data to be
automatically identify drills in addition to velocity.

------------------------------------------------------------------------

## 8. Elaborate — derive custom metrics

`elaborate()` is a `dplyr::mutate()` wrapper that is safe to use on
`motion_trace` objects — it preserves the `metadata` and `quality`
attributes that a bare `mutate()` would silently drop.

Use it any time you want to add a computed column before allocating or
quantitating. Here we create **curved load** - a metric I ... Just made
up for the sake of demonstrating this feature.

```{r elaborate}
trace <- trace |>
  elaborate(curved_load = velocity * abs(angular_velocity))
```

------------------------------------------------------------------------

## 9. Allocate — assign intensity bands

`allocate()` cuts each derivative into labelled bands using
user-supplied thresholds. You can call it multiple times with different
`allocation_name` values to layer multiple profiling schemes. Here, we
specify the bands in the script itself. But you can also call them
through an external file (outlined outside of quickstart)

```{r allocate}
trace <- trace |>
  allocate(allocation = list(
      velocity = c(0, 2, 4, 6, 8),
      acceleration = c(0, 1, 2, 3, 4),
      angular_velocity = c(0,90,180)),
    allocation_name = "gps_standard"
  )

```

------------------------------------------------------------------------

## 10. Quantitate — summarise by segment and intensity band

`quantitate()` aggregates the motion trace by designation segment and
intensity band, returning a tidy summary tibble. The full quality log is
preserved as an attribute on the result. You can calculate any basic
aggregate parameter here. You'd usually want to do this by
designation....

```{r quantitate}
summary_tbl <- trace |>
  quantitate(
    allocation = "gps_standard",
    derivative = "velocity",
    scope = "designation",
    active_only = FALSE,
    n_frames = dplyr::n(),
    time_s = dplyr::n() / 10,
    dist_m = sum(distance, na.rm = TRUE),
    mean_vel = mean(velocity, na.rm = TRUE)
  )

summary_tbl
```

But you can also get session totals. Here you'd probably want to keep
active only, however.

```{r quantitate-session}
session_totals <- trace |>
  quantitate(allocation = "gps_standard",
    derivative = "velocity",
    scope = "session",
    active_only = TRUE,
    time_s = dplyr::n() / 10,
    dist_m = sum(distance, na.rm = TRUE),
    peak_vel = max(velocity, na.rm = TRUE))

session_totals
```

------------------------------------------------------------------------

## Appendix -  Inspect the quality log

Let's take a look at the quality log I outlined earlier now that we are
at the end. Notice how it preserved everything we did to the trace?

```{r quality-report}
quality_report(trace)
```

------------------------------------------------------------------------

## Saving Results

R attributes (including the quality log) are not preserved by write_csv.
Use `saveRDS()` to keep the full object, or extract the quality log
before writing to CSV:

```{r save, eval=FALSE}
# Preserve the full motion_trace with quality log
saveRDS(trace, "session_processed.rds")

# Export summary to CSV (attributes are dropped — this is expected)
readr::write_csv(summary_tbl, "session_summary.csv")

# Save quality log separately if you need it alongside the CSV
qual_log <- attr(trace, "quality")
saveRDS(qual_log, "session_quality_log.rds")
```

------------------------------------------------------------------------

## Pipeline at a glance

The real value in *motiongrammar* is building pipes. We took it step by
step above, but below is a very real pipeline of how this package can be
utilised.

```{r}
trace <- initiate(session = '~/motiongrammar/data/example_gps_file_short.csv',
                  method = 'guess') |>
  coordinate(from = "latlong", 
             to = "xyz", 
             norm = TRUE) |>
  interpolate(hz = 10) |>
  filtrate(method = "butterworth", 
           target = "coordinates", 
           cutoff = 1.5) |>
  derivate() |>
  designate() |>
  allocate(allocation_name = "gps_standard") |>
  quantitate(allocation = "gps_standard",
    derivative = "velocity",
    scope = "session",
    time_s = dplyr::n() / 10,
    dist_m = sum(distance, na.rm = TRUE))
```
