---
title: "motiongrammar Quickstart"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{motiongrammar Quickstart}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  warning = FALSE,
  message = FALSE
)
```

*motiongrammar* brings GPS, RFID, and optical tracking data into a
single tidyverse-style pipeline. Within one chain of commands, we can
transform our raw tracking data into aggregate summary tables. This
includes transforming from (latitude,longitude) to (x,y), applying
filters to attenuate noise, calculating custom time-series, and applying
bands and identifying periods. Along the way, *motiongrammar* keeps a
log of everything you have done to the dataset to maximise
reproducibility.

```{r pipeline-hook, eval=FALSE}
library(motiongrammar)
library(tidyverse)

exemplar_trace <- initiate(
    session  = "gps_session.csv",
    template = load_csv_template("my_device.xml")
  ) |>
  coordinate(norm = TRUE) |>
  interpolate(hz = 10, max_gap_frames = 20) |>
  filtrate(method = "butterworth", target = "coordinates", cutoff = 0.4) |>
  derivate(derivatives = c("velocity", "acceleration", "angular_velocity")) |>
  designate(method = "manual", file = "drill_labels.csv") |>
  elaborate(velocity = pmin(velocity, 8),
            angular_velocity = pmin(abs(angular_velocity), 360),
            curved_load = velocity * angular_velocity) |>
  allocate_csv(file = "intensity_bands.csv")
```

That's the complete pipeline. The rest of this guide walks through each
step with a real GPS session, explaining what each function does, how to
tune it, and what to watch out for.

------------------------------------------------------------------------

```{r load, include=FALSE}
library(motiongrammar)
library(tidyverse)

path <- system.file("data", "marquee_gps_file.csv", package = "motiongrammar")
```

------------------------------------------------------------------------

## A. Guess — let motiongrammar figure out the file type

`initiate(source = "guess_csv")` takes it's best shot at working out our
csv format automatically. We would not use this in a pipeline as it is
not standardised. But it is useful to quickly identify column names
using fuzzy logic, and handles issues we sometimes see with vendors with
empty rows/rows that need to be skipped etc.

```{r initiate-guess}
exemplar_trace <- initiate(source = "guess_csv", 
               session = path, 
               verbose = FALSE)

exemplar_trace
```

A key feature of motiongrammar is the *quality report*. It:

1.  Keeps trace of every transformation you do on the data, whether it
    be a calculation, filter or coordinate change (important in the
    academic world and for debugging!)
2.  Calls out anything that may be an error or issue in the data set.
    Quality errors are not always glaring issues that need to be
    addressed. They are just nudges that something may be amiss.

The quality report updates automatically at every step. Later on, we
will look at how to examine quality for a single step only.

```{r quality-report-1}
quality_report(exemplar_trace)
```

> The 14-second temporal gap and duplicate timestamps are real features
> of this export. The quality system catches them automatically.

------------------------------------------------------------------------

## B. Inspect the detected template

`guess_csv` only imports the four core columns (time, lat, lng,
altitude). In the real world, you may want to keep vendor specific
series' like metabolic power or player load `guess_csv_template()` shows
exactly what was detected and allows us to extend it.

```{r guess-template}
my_csv_template <- guess_csv_template(path)
my_csv_template
```

Add the device-specific extra columns, then save the template. Once
saved, you can import a device format consistently. As a result, the
prior inspection and this step are ideally a one-off. Vendors sometimes
change formats, so in your own pipeline, it is important to keep
everything you have set to identify where changes are coming from.

NB: in camera systems, you may use frame as unix time. This is totally
fine.

```{r build-template}
my_csv_template <- csv_template(
  col_unix = "UnixTimestamp",
  col_lat = "Latitude",
  col_lng = "Longitude",
  col_altitude = "Elevation",
  col_extra = c("HeartRate", "PlayerLoad", "MetabolicPower", "Cadence"),
  skip = 6L
)

template_path <- file.path(tempdir(), "marquee_gps_template.xml")
save_csv_template(my_csv_template, template_path)
```

Reload at any time (from another script, a different machine, years
later). After your first pass, *this* should be how you initiate a
motiongrammar pipeline.

```{r reload-template}
my_csv_template <- load_csv_template(template_path)
```

------------------------------------------------------------------------

## 1. Initiate — import with the template

```{r initiate-template}
exemplar_trace <- initiate(session = path, 
                           template = my_csv_template,
                           verbose = FALSE)
exemplar_trace
```

HeartRate, PlayerLoad, MetabolicPower, and Cadence are now columns that
will carry through every downstream step.

------------------------------------------------------------------------

## 2. Coordinate — project to metres

GPS coordinates are spherical (degrees). `coordinate()` projects them to
a flat Cartesian plane so distances, velocities, and headings are in
metres. `norm = TRUE` places the first position at the origin. Check out
the documentation on coordinate to see what your other options are.
motiongrammar has methods to handle automatic and manual origins.

```{r coordinate}
exemplar_trace <- coordinate(exemplar_trace, norm = TRUE)
```

Let's plot what we have so far. In the figure below, we can see that the
player returned to the locker room or GPS signal dropped out causing a
spike. This is a very common scenario, and something we will handle in
this quickstart.

```{r plot-raw-path}
ggplot(exemplar_trace, aes(x, y)) +
  geom_path(alpha = 0.6, 
            linewidth = 0.4) +
  coord_fixed() +
  labs(title = "Projected trajectory (before interpolation)",
       x = "x (m)", y = "y (m)") +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

------------------------------------------------------------------------

## 3. Interpolate — regularise the frame rate

GPS signals drop out. `interpolate()` builds a regular 10 Hz time grid
and fills short gaps linearly. Gaps longer than `max_gap_frames` rows
are left as `NA` rather than invented — preserving the 14-second gap
flagged earlier.

Advanced users may be interested in the other methods offered here,
albeit, they are less mainstream.

```{r interpolate}
exemplar_trace <- interpolate(exemplar_trace, 
                              hz = 10, 
                              method = "linear", 
                              max_gap_frames = 20)
```

------------------------------------------------------------------------

## 4. Filtrate — smooth the coordinates

Your vendor has likely filtered your data for you, so you may not even
need this step. If you get a tiny cut off (less than 0.5), it's likely
your work has already been done for you. In saying that, filtering may
be necessary and is an important skill as a biomechanist, so it's
included here.

### Find the cutoff

We will go into the theory behind this in a later text. But for the time
being, you just need to know the following:

1.  A low pass butterworth filter is what is usually used to smooth
    human movement data.
2.  The goal is to attenuate (or weaken) noise, not abruptly remove it.
3.  We need to identify a cut-off that marks the trade off between
    attenuating the noise in a signal, without removing real motion
    data.

`filtrate_cutoff()` sweeps a range of Butterworth cutoffs and plots the
residual curve. The elbow marks where smoothing starts removing real
signal rather than noise.

```{r filtrate-cutoff, fig.height=5}
filtrate_cutoff(
  exemplar_trace,
  method = "butterworth",
  target = "coordinates",
  axis = "x",
  plot_type = "both",
  cutoff_range = c(0.1, 4),
  n_steps = 40
)
```

A cutoff of **\~1–2 Hz** is typical for GPS position data at 10 Hz. The
dashed red line is the suggested value — adjust based on the PSD panel.

We can tell this data is already filtered because the filter is very
small. Another indicator that filtering has adequetely been performed by
the vendor, is seeing more of a "straight line" than an elbow.

For the purpose of educating, I'm going to utilise the suggested 0.4 cut
off. In the real world, I'd probably just work off unfiltered
coordinates (which motiongrammar also supports!)

### Apply the filter

`filtrate()` stores the smoothed signal in `f_x`, `f_y` columns and
leaves the originals untouched. Future steps will use the filtered
coordinates by default, but we can change this.

You'll also note that there is a "target" argument here. Whether to
filter coordinates (as we have done) or derivatives (velocity, angular
velocity and acceleration) is a bit of a debate in biomechanics. Given
that coordinates are more common (and library only one filter, versus
separate cut offs for angular velocity, acceleration and velocity)
that's what I typically work off.

```{r filtrate}
exemplar_trace <- filtrate(exemplar_trace, 
                           method = "butterworth", 
                           target = "coordinates",
                           cutoff = 0.4)
```

Let's visualise what we have done so far:

```{r plot-filtered}
exemplar_trace |> 
drop_na(f_x) |>
  ggplot(aes(x = unix_time)) +
  geom_line(aes(y = x,
                colour = "Raw"),
            linewidth = 0.3) +
  geom_line(aes(y = f_x, colour = "Filtered"), 
            linewidth = 0.5) +
  scale_colour_manual(values = c(Raw = "#D55E00", 
                                 Filtered = "#0072B2")) +
  labs(title = "x coordinate: raw vs filtered",
       x = "Unix time (s)", y = "x (m)", colour = NULL) +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

We see some minor differences, but overall the small cut off led to a
very forgiving filter.

------------------------------------------------------------------------

## 5. Derivate — velocity, acceleration, angular velocity

`derivate()` computes derivatives using central differences. It uses the
filtered `f_x`, `f_y` columns by default.

```{r derivate}
exemplar_trace <- derivate(
  exemplar_trace,
  use_filtered = TRUE,
  derivatives = c("velocity", "acceleration", "angular_velocity")
)
```

You are probably familiar with velocity and acceleration. Angular
velocity captures whether the player is running straight or turning as a
continuous signal. Hats off to *Alice Sweeting* and co for her use of
this in netball (we should all be using this more ;))

```{r plot-angular-velocity}
drop_na(exemplar_trace, angular_velocity) |>
  ggplot(aes(f_x, f_y, colour = angular_velocity)) +
  geom_path(linewidth = 1) +
  coord_fixed() +
  scale_colour_viridis_c(option = "viridis", begin = 0.1, end = 0.9) +
  labs(title = "Angular velocity mapped onto trajectory",
       x = "x (m)", y = "y (m)", colour = "deg/s") +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

> **GPS artefact — player off field.** The extreme angular velocity
> values visible at one edge of the trajectory are not biomechanical.
> The 14-second gap flagged by `initiate` in step 1 corresponds to the
> player leaving the field (rotation, changing a shirt). GPS continued
> recording at an off-field position, then snapped back when the player
> returned — producing a position jump that `derivate()` reads as
> extreme velocity and angular velocity. This is a real-world data
> problem that the quality log flags but cannot automatically clean; we
> address it explicitly in step 9.

------------------------------------------------------------------------

## 6. Designate — identify activity bouts

### 8a. Automated segmentation — and generating a skeleton

`designate()` uses PELT changepoint detection on the velocity signal to
split the session into contiguous active bouts and relief periods.
`v_threshold` tells the algorithm what counts as movement.

```{r designate-corbett}
corbett_trace <- designate(exemplar_trace,
                           method = "corbett",
                           v_threshold = 2)
```

```{r plot-designate}
corbett_trace |>
  filter(velocity < 10) |>
  ggplot(aes(x = unix_time, y = velocity, colour = designation)) +
  geom_line(linewidth = 0.4) +
  labs(title = "Velocity coloured by automated segment",
       x = "Time (s)", y = "Velocity (m/s)", colour = NULL) +
  theme_minimal() +
  theme(panel.grid = element_blank(), legend.position = "bottom")
```

The automated labels (`active_1`, `relief_1` etc.) are generic. For
structured sessions you will want named drill periods. Use
`export_designations()` to write the detected ranges to a CSV — this is
your skeleton to edit:

```{r export-designations}
skeleton_path <- tempfile(fileext = ".csv")
export_designations(corbett_trace, path = skeleton_path)
read.csv(skeleton_path)
```

Two edits you will typically make before saving this as your session
label file:

-   **Rename** `active_1`, `relief_1` etc. to meaningful drill labels
    (`warmup_mobility`, `tempo_run`, `change_of_direction`, etc.)
-   **Shift timestamps by a few seconds** where needed — common when
    your stopwatch or video system is slightly offset from GPS time, or
    where you want to trim the edges of a period

You will notice the algorithm did a *reasonable job*. But in the real
world, we would use 8b since the efforts being clipped is not ideal.

### 8b. Importing manual designations (preferred!)

In practice the label file usually comes from a coaching system, a video
coder, or an edited version of the skeleton above. The expected format
is three columns: `designation`, `unix_start`, `unix_end`.

```{r read-desig-csv}
desig_csv <- system.file("data", "motiongrammar_quickstart_designations.csv", package = "motiongrammar")
```

```{r designate-manual}
exemplar_trace <- designate(
  exemplar_trace,
  method = "manual",
  file = desig_csv
)
quality_report(exemplar_trace, step = 'designate')

```

The quality log records the source file path and every row assigned,
making the designation fully auditable. If your column names differ,
`cols = 'guess'` handles fuzzy matching, or supply an explicit mapping:

```{r designate-cols-options, eval=FALSE}
# Fuzzy detection — tolerates "label", "start_time", "stop" etc.
designate(exemplar_trace, 
          method = "manual",
          file = desig_csv, 
          cols = "guess")

# Explicit mapping — most robust for automated pipelines
designate(exemplar_trace, method = "manual",
          file = desig_csv,
          cols = c(designation = "designation",
                   unix_start = "unix_start",
                   unix_end = "unix_end"))
```

------------------------------------------------------------------------

## 7. Elaborate (optional) — add custom time-series

`elaborate()` is a `motion_trace`-safe `dplyr::mutate()` wrapper. It
preserves the `metadata` and `quality` attributes that a bare `mutate()`
would silently drop. In our example below, we are going to manually
remove the pesky velocity spike. Whilst not scientifically ideal, it is
common for practitioners to have a threshold for inhuman velocity (I
know some AFL teams use 32 km/hr for instance)

### Removing the off-field artefact

The position jump from the player leaving the field produced:

-   a velocity spike well above any realistic sprinting speed (\>8.5 m/s
    ≈ 31 km/h here.)
-   an angular velocity spike of hundreds or thousands of deg/s —
    impossible biomechanically, caused by the GPS unit snapping back to
    a new position in a single frame

Here, we apply a simple clamp.

Setting `log_expr = TRUE` tells motiongrammar to record the exact
expression used for each column — critical when you are overwriting a
reserved pipeline column like `velocity`, because downstream steps
(allocate, quantitate) will silently use the new values. Without an
audit trail you lose traceability of what your reported numbers actually
represent.

```{r elaborate}
exemplar_trace <- elaborate(
  exemplar_trace,
  velocity = pmin(velocity, 8.5),                # hard ceiling: ~28.8 km/h
  angular_velocity = pmin(abs(angular_velocity), 360), # pivot ≈ 360 deg/s max
  curved_load = velocity * angular_velocity,      # direction-weighted load (clean values)
  log_expr = TRUE
)
```

```{r elaborate-quality}
quality_report(exemplar_trace, step = 'elaborate')
```

In the report above, the `⚠ Overwrote` line flags that `velocity` and
`angular_velocity` are reserved columns — values computed by
`derivate()` that the rest of the pipeline depends on. The `Expressions`
block records *exactly what you replaced them with*, so any colleague
(or future you) can reconstruct what the output numbers mean without
digging through script history.

Elaborate allows you to calculate anything you might want on a
time-series level. That might be *moving averages* (initially proposed
by Varley et al., coined and popularised as *peak demands* by Delaney et
al.)

------------------------------------------------------------------------

## 8. Allocate — assign intensity bands

`allocate_csv()` reads a band definition file and applies every profile
it contains in a single call. This is the most common workflow in
practice — your team or lab maintains one canonical thresholds file that
everyone imports, keeping analyses comparable across sessions and
analysts.

```{r read-alloc-csv}
alloc_csv <- system.file("data", "motiongrammar_quickstart_allocations.csv", package = "motiongrammar")
```

```{r allocate}
exemplar_trace <- allocate_csv(exemplar_trace, 
                               file = alloc_csv)
```

The CSV defines two profiles (`gps_standard` and `quickstart_simple`),
each applied in one call. Reference them by name in `quantitate()`.

If you need a quick one-off scheme without a file, `allocate()` accepts
a named list directly. This is useful for exploratory work or computed
thresholds from a model (e.g. Sweeting's clustering approach):

```{r allocate-manual-example, eval=FALSE}
exemplar_trace <- allocate(
  exemplar_trace,
  allocation = list(
    velocity = c(0, 1.5, 3.5, 5.0, Inf),
    acceleration = c(0, 1.0, 2.0, Inf)
  ),
  allocation_name = "custom_session"
)
```

------------------------------------------------------------------------

## 9. Quantitate — summarise

`quantitate()` aggregates by designation segment and intensity band. The
full quality log is preserved as an attribute on the result. Call it
once per derivative you want to profile — the results are separate
tibbles that can be joined on `designation` downstream.

### Velocity profile by segment

```{r quantitate-velocity}
by_vel <- quantitate(
  exemplar_trace,
  allocation = "gps_standard",
  derivative = "velocity",
  scope = "designation",
  active_only = TRUE,
  time_s = n() / 10,
  peak_vel = max(velocity, na.rm = TRUE),
  mean_hr = mean(HeartRate, na.rm = TRUE)
)

by_vel
```

### Acceleration profile by segment

```{r quantitate-acceleration}
by_accel <- quantitate(
  exemplar_trace,
  allocation = "gps_standard",
  derivative = "acceleration",
  scope = "designation",
  active_only = TRUE,
  time_s = n() / 10,
  peak_accel = max(acceleration, na.rm = TRUE),
  mean_accel = mean(acceleration, na.rm = TRUE),
  n_efforts = sum(acceleration >= 2, na.rm = TRUE)  # frames at high+ accel
)

by_accel
```

### Angular velocity (turning) profile by segment

```{r quantitate-angular}
by_ang <- quantitate(
  exemplar_trace,
  allocation = "gps_standard",
  derivative = "angular_velocity",
  scope = "designation",
  active_only = TRUE,
  time_s = n() / 10,
  peak_ang_vel = max(angular_velocity, na.rm = TRUE),
  mean_ang_vel = mean(angular_velocity, na.rm = TRUE),
  n_cod = sum(angular_velocity >= 90, na.rm = TRUE)  # frames in turn+
)

by_ang
```

### Session totals — all three derivatives

```{r quantitate-session}
session_totals <- quantitate(
  exemplar_trace,
  allocation = "gps_standard",
  derivative = "velocity",
  scope = "session",
  active_only = TRUE,
  time_s = n() / 10,
  peak_vel = max(velocity, na.rm = TRUE),
  mean_hr = mean(HeartRate, na.rm = TRUE),
  peak_accel = max(acceleration, na.rm = TRUE),
  mean_ang_vel = mean(angular_velocity, na.rm = TRUE),
  peak_ang_vel = max(angular_velocity, na.rm = TRUE),
  curved_load = sum(curved_load, na.rm = TRUE)
)

session_totals
```

------------------------------------------------------------------------

## Appendix: quality log

Every step appends to the quality log. Inspect it at any point — here at
the end you can see the full provenance of everything done to the trace:

```{r quality-final}
quality_report(exemplar_trace)
```

------------------------------------------------------------------------

## Saving results

`write_csv` drops R attributes. Use `saveRDS` to keep the full object:

```{r save, eval=FALSE}
saveRDS(exemplar_trace, "session_processed.rds")

## Save our summary files
readr::write_csv(by_vel,"by_velocity.csv")
readr::write_csv(by_accel,"by_acceleration.csv")
readr::write_csv(by_ang,"by_angular_velocity.csv")

## Save quality log as it's own file.
saveRDS(attr(exemplar_trace, "quality"), "quality_log.rds")
```
